---
title: "Fixed-Duration Tunnel States with `treatseqr`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fixed-Duration Tunnel States with `treatseqr`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(treatseqr)
```

# Introduction

This vignette demonstrates a more complex model structure using `treatseqr`,
combining fixed-duration tunnel states with full-length tunnel states in the
same model.

The scenario is an oncology treatment sequence in which patients on first-line
therapy may enter a surgical intervention. The surgery tunnel has a fixed
duration of 3 cycles, after which patients either respond (entering a
full-length responder tunnel with good outcomes) or do not respond (entering a
second fixed-duration tunnel representing a period of poor post-surgical
recovery). Patients who do not respond face elevated mortality and may
accumulate in a **holding state** at the end of the non-responder tunnel.
Finally, at any point patients may elect to forgo further active treatment and
enter a palliative full-length tunnel with poor outcomes.

The five model states are:

| State | Type | Tunnel length |
|---|---|---|
| `on_trt` | Pre-tunnel | — |
| `surgery` | Fixed-duration tunnel | 3 cycles |
| `failed_surg` | Fixed-duration tunnel (holding state) | 2 cycles |
| `responder` | Full-length tunnel | `th` cycles |
| `palliative` | Full-length tunnel | `th` cycles |

Because tunnels 1 and 2 are short and tunnels 3 and 4 span the full model
horizon, this model requires the variable-length `tunnel_lengths` argument.

# Model specification

We set the time horizon to 20 cycles and define the tunnel lengths individually:

```{r specify_m}
th <- 20L

m_spec <- specify_m(
  tunnel_lengths    = c(3L, 2L, th, th),
  pre_tunnel_states = 1L
)

m_spec$matrix_size
```

The total matrix size is `r m_spec$matrix_size`: 1 pre-tunnel state + 3 + 2 +
20 + 20 tunnel cycles + 1 death state.

# Transition probabilities

The transition probability source list must follow the standard `treatseqr`
conventions: states listed in sequence order; each state contains transition
probability vectors to all other states (or `NULL`/omit for structural zeros).
With variable-length tunnels, each vector must match the length of the
**source** state's tunnel (or the pre-tunnel horizon `th` for `on_trt`).

Just to make things very simple, we just use some very simple fixed transitions
this time, but these could be made time-dependent here without any changes
required in the rest of the code:

```{r define_tps}
tp_source <- list(
  # ----------------------------------------------------------------
  # on_trt (pre-tunnel): vectors of length th
  # ----------------------------------------------------------------
  on_trt = list(
    surgery     = seq(0.05, 0.02, length.out = th),
    failed_surg = NULL,
    responder   = NULL,
    palliative  = rep(0.02, th),
    die         = rep(0.01, th)
  ),

  # ----------------------------------------------------------------
  # surgery (fixed-duration, 3 cycles):
  #   d = 1, 2 : patients recover — low exit probability
  #   d = 3    : decision-tree node — p_stay = 0, all probability
  #              allocated to response, non-response, or death
  # ----------------------------------------------------------------
  surgery = list(
    failed_surg = c(0, 0, 0.35),
    responder   = c(0, 0, 0.55),
    palliative  = c(0.01, 0.01, 0),
    die         = c(0.04, 0.04, 0.10)
  ),

  # ----------------------------------------------------------------
  # failed_surg (fixed-duration, 2 cycles):
  #   d = 1 : elevated mortality; patients may opt to palliative
  #   d = 2 : even higher mortality; p_stay = 0.70 (holding state)
  #           — patients who have not yet exited accumulate here
  #           with the same d = 2 mortality applied every subsequent
  #           cycle
  # ----------------------------------------------------------------
  failed_surg = list(
    responder  = c(0, 0),
    palliative = c(0.10, 0.10),
    die        = c(0.15, 0.20)
  ),

  # ----------------------------------------------------------------
  # responder (full-length, th cycles): good outcomes
  # ----------------------------------------------------------------
  responder = list(
    palliative = seq(0.02, 0.05, length.out = th),
    die        = seq(0.01, 0.05, length.out = th)
  ),

  # ----------------------------------------------------------------
  # palliative (full-length, th cycles): poor outcomes
  # ----------------------------------------------------------------
  palliative = list(
    die = rep(0.25, th)
  )
)
```

# Validating transition probabilities

Before populating the matrix we validate the source list. Passing
`tunnel_lengths` tells the validator which vector length to expect for each
tunnel state.

```{r validate}
valid_tp <- validate_tp_source(
  tp_source      = tp_source,
  tunnel_lengths = c(3L, 2L, th, th)
)

lobstr::tree(
  rapply(valid_tp, function(x) round(x, 3), how = "list")
)
```

The NULL entries for `on_trt$failed_surg` and `on_trt$responder` have been
replaced with zero vectors of length th, and `failed_surg$responder` filled with
zeros of length 2. In addition, the validator expands every state's destination
set to the full canonical set — backwards slots (e.g. `surgery$on_trt`,
`responder$surgery`) are auto-added as zero vectors.

# The decision-tree node at the end of `surgery`

At `d = 3` — the last cycle of the surgery tunnel — the transition probability
vectors are constructed so that `p_stay = 0`:

```{r decision_node}
# Probabilities out of surgery at each tunnel cycle d (rows = destinations):
surgery_exit <- do.call(rbind, lapply(valid_tp$surgery, identity))
rownames(surgery_exit) <- names(valid_tp$surgery)
round(t(surgery_exit), 3)
```

Column sums confirm that `d = 3` is a true decision-tree node (sums to 1,
leaving no residual probability of remaining in the surgery tunnel):

```{r decision_node_check}
colSums(round(t(surgery_exit), 3))
```

This pattern — zeroing `p_stay` at the last tunnel cycle and distributing all
remaining probability across destination states — is the recommended approach
for fixed-duration tunnel states with a defined exit point.

# The holding state at the end of `failed_surg`

The `failed_surg` tunnel illustrates the opposite design choice. At `d = 2`
(the last defined cycle), a substantial `p_stay` remains:

```{r holding_state}
# Probabilities out of failed_surg at each tunnel cycle d:
failed_exit <- do.call(rbind, lapply(valid_tp$failed_surg, identity))
rownames(failed_exit) <- names(valid_tp$failed_surg)
round(t(failed_exit), 3)
```

```{r holding_state_pstay}
# Implied p_stay at each cycle of failed_surg:
1 - colSums(round(t(failed_exit), 3))
```

Patients who have not exited `failed_surg` by `d = 2` remain there until going
to palliative care or death, with the `d = 2` transition probabilities
re-applied each subsequent cycle. This is conceptually equivalent to an
exponential extrapolation in the holding state — the same probability of dying
(`0.20`) or transitioning to palliative (`0.10`) is applied every cycle to
whoever remains. This is an intentional modelling choice which a modeller may
want to employ. Note that instead of this, one could simply add a new tunnel and
introduce a decision-tree node at `d = 2` to force all patients out of
`failed_surg` at that point. The choice depends on the clinical scenario and
modelling goals. The benefit of this choice is it reduces the size of `M`,
improving computational efficiency, but it does simplify the disease modelling.

If instead patients should be guaranteed to exit at `d = 2`, set
`die[2] + palliative[2] = 1` and `p_stay = 0`, following the same
decision-tree node pattern used for `surgery`. Then, Kaplan-Meier for surgical
non-response outcomes could be used to power what happens to those patients in
the model.

# Generating M and extrapolating

```{r generate_and_extrapolate}
m <- generate_m_list(m_specification = m_spec, transition_prob_list = valid_tp)
trace <- extrapolate_treatseqr(m = m, spec = m_spec)
```

A quick sanity check — column sums of the trace should equal 1 throughout
(all population accounted for):

```{r check_trace}
colSums(trace)
```

# Consolidation and outputs

```{r consolidate}
res <- consolidate_treatseqr_trace(
  full_trace  = trace,
  spec        = m_spec,
  state_names = m$state_names
)
```

## State populations over time (`t`)

```{r plot_t}
graphics::matplot(
  res$t,
  type = "l", lty = 1,
  xlab = "Cycle", ylab = "Population",
  main = "State populations over absolute time"
)
graphics::legend(
  "topright",
  legend = colnames(res$t),
  col    = seq_len(ncol(res$t)),
  lty    = 1,
  cex    = 0.8
)
```

The `failed_surg` population decays slowly compared with `surgery` because of
the holding state — patients continue to accumulate at `d = 2` across many
cycles before eventually dying or transitioning to palliative.

## Within-tunnel survival (`d`) no longer works in this setting!

As mentioned in the documentation, the `d` output is only meaningful when
patients cannot accumulate inside of one `d` for one tunnel state. In this
"holding pattern" conceptual model (which is somewhat of an edge-case), you can
no longer derive within-tunnel survival curves because the `d` dimension is no
longer a pure measure of time spent in the state. The same is also technically
true (but less pronounced) in the setting whereby a patient can loop back into a
previous tunnel state (e.g., a relapse followed by non-progression back to the
previous tunnel).

```{r plot_d}
graphics::matplot(
  res$d,
  type = "l", lty = 1,
  xlab = "Cycles since tunnel entry", ylab = "Proportion remaining",
  main = "Within-tunnel survival"
)
graphics::legend(
  "topright",
  legend = colnames(res$d),
  col    = seq_len(ncol(res$d)),
  lty    = 1,
  cex    = 0.8
)
```

The `surgery` within-tunnel survival curve drops sharply to zero at `d = 3`
(the decision-tree node forces everyone out), whereas `failed_surg` shows a
much slower tail as patients accumulate in the holding state at `d = 2`. The
full-length tunnels (`responder`, `palliative`) display smoother declines
reflecting their per-cycle mortality rates.

Because this structure is a **strictly progressive** structure not allowing
patients to move back to previous tunnels, the only violation here is
`failed_surg`, because patients are accumulating at `d = 2` and thus `d` is no
longer a pure measure of time spent in the state. In a more complex model with
loops back to previous tunnels, multiple states could violate this assumption,
but the violation would be harder to see. this is because it may be much more
gradual and smooth, depending on the probability each cycle for patients to loop
back to a previous tunnel.

## Expected time in state

The total expected cycles spent in each state across the cohort:

```{r expected_time}
round(colSums(res$t), 2)
```
