---
title: "Backwards Transitions in Relapsing-Remitting Disease"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Backwards Transitions in Relapsing-Remitting Disease}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(treatseqr)
```

# Introduction

This vignette demonstrates backwards transitions in `treatseqr` using a
relapsing-remitting multiple sclerosis (RRMS) setting. In contrast to oncology
treatment-sequence models — where disease progression is typically assumed
irreversible — RRMS patients may experience relapses that resolve without
accumulating permanent disability, returning to an earlier disease state. This
is precisely the backwards-transition mechanism built into `treatseqr`.

The model represents escalating levels of disease burden as a sequence of tunnel
states. A brief model-entry pre-tunnel state (`newly_dx`) routes patients into
`stage_1` — the minimal-disability, between-relapse state — which is itself
modelled as a full-length tunnel. Modelling `stage_1` as a tunnel (rather than
a pre-tunnel state) is what allows the backwards-transition mechanism to work: in
`treatseqr`, transitions from within a tunnel back to an earlier state can only
target other **tunnel** start positions, so any state that needs to *receive*
backwards transitions must itself be a tunnel state.

An additional benefit is clinical richness: the `d` dimension of `stage_1`
tracks how long a patient has been continuously in the between-relapse state
since their last relapse resolution. Patients who resolve from `stage_2` back to
`stage_1` re-enter at `d = 1`, resetting this counter. To go beyond this point
of sophistication essentially requires unlimited dimensionality to the
transition probability matrix. In such cases (e.g., where a counter of how many
times individuals have been to each state and for how long is needed), a
microsimulation may be more appropriate than any kind of cohort model could ever
realistically be.

During each relapse tunnel, patients may:

- **Resolve** (backwards transition) — return to the start of the previous
  tunnel at any cycle `d = 1, 2, …, r_len`, not just at the end.
- **Progress** (forward transition) — advance to the next relapse level, again
  possible at any cycle during the relapse.
- **Die** — small background mortality throughout.
- **Remain in the relapse** — `p_stay > 0` at `d = 1` and `d = 2`; a
  decision-tree node at `d = r_len` forces a clean exit for any patients who
  have not yet resolved or progressed.

Patients who reach the most severe stage enter a **secondary progressive**
tunnel with no possibility of reversal, illustrating how `treatseqr` accommodates
both reversible and irreversible disease within the same model.

The five model states are:

| State | Type | Tunnel length | Clinical meaning |
|---|---|---|---|
| `newly_dx` | Pre-tunnel | — | Model entry at RRMS diagnosis |
| `stage_1` | Full-length tunnel | `th` | Minimal disability (EDSS 0–2); between relapses |
| `stage_2` | Fixed-duration (DTN) | `r_len` | Active relapse, moderate disability (EDSS 3–4) |
| `stage_3` | Fixed-duration (DTN) | `r_len` | Active relapse, severe disability (EDSS 5–6) |
| `stage_4` | Full-length tunnel | `th` | Secondary progressive phase (EDSS 7+); irreversible |

# Model specification

```{r specify_m}
r_len <- 3L
th <- 40L

m_spec <- specify_m(
  tunnel_lengths    = c(th, r_len, r_len, th),
  pre_tunnel_states = 1L
)

m_spec$matrix_size
```

The total matrix size is `r m_spec$matrix_size`: 1 pre-tunnel state + `r th` +
3 + 3 + `r th` tunnel cycles + 1 death state.

# Transition probabilities

Backwards and forward transitions within the relapse tunnels are non-zero at
every `d`, rising toward the decision-tree node at `d = r_len`. Destinations
not listed are `NULL` and will be auto-filled as zero vectors by the validator.

```{r define_tps}
tp_source <- list(
  # ------------------------------------------------------------------
  # newly_dx (pre-tunnel): vectors of length th
  #   nearly all patients enter stage_1 immediately after diagnosis;
  #   small background mortality
  # ------------------------------------------------------------------
  newly_dx = list(
    stage_1 = rep(0.97, th),
    stage_2 = NULL,
    stage_3 = NULL,
    stage_4 = NULL,
    die     = rep(0.01, th)
    # implied p_stay = 0.02 every cycle
  ),

  # ------------------------------------------------------------------
  # stage_1 (full-length, th cycles): minimal disability
  #   low but ongoing relapse risk each cycle
  # ------------------------------------------------------------------
  stage_1 = list(
    stage_2 = rep(0.06, th),
    stage_3 = NULL,
    stage_4 = NULL,
    die     = rep(0.01, th)
    # implied p_stay = 0.93 every cycle
  ),

  # ------------------------------------------------------------------
  # stage_2 (fixed-duration, r_len cycles):
  #   d = 1, 2 : resolution or progression possible but not certain
  #              — p_stay > 0, most patients remain in the relapse
  #   d = 3    : decision-tree node — p_stay = 0, all probability
  #              allocated to resolution, progression, or death
  #
  #   stage_1 (backwards): relapse resolves without disability gain
  #   stage_3 (forward):   relapse progresses to higher disability
  # ------------------------------------------------------------------
  stage_2 = list(
    stage_1 = c(0.15, 0.20, 0.50), # backwards: rises each cycle
    stage_3 = c(0.05, 0.08, 0.40), # forward:   rises each cycle
    stage_4 = NULL,
    die     = c(0.02, 0.02, 0.10)
    # d=1 implied p_stay = 0.78
    # d=2 implied p_stay = 0.70
    # d=3 implied p_stay = 0.00  (decision-tree node)
  ),

  # ------------------------------------------------------------------
  # stage_3 (fixed-duration, r_len cycles):
  #   d = 1, 2 : partial resolution or progression possible
  #   d = 3    : decision-tree node — p_stay = 0
  #
  #   stage_2 (backwards): partial resolution; returns to moderate relapse
  #   stage_4 (forward):   enters secondary progressive phase
  # ------------------------------------------------------------------
  stage_3 = list(
    stage_2 = c(0.10, 0.15, 0.35), # backwards: rises each cycle
    stage_4 = c(0.03, 0.08, 0.50), # forward:   rises each cycle
    die     = c(0.03, 0.03, 0.15)
    # d=1 implied p_stay = 0.84
    # d=2 implied p_stay = 0.74
    # d=3 implied p_stay = 0.00  (decision-tree node)
  ),

  # ------------------------------------------------------------------
  # stage_4 (full-length, th cycles): secondary progressive — no reversal
  #   rising mortality reflects continued disease progression
  # ------------------------------------------------------------------
  stage_4 = list(
    die = seq(0.05, 0.20, length.out = th)
  )
)
```

# Validating transition probabilities

```{r validate}
valid_tp <- validate_tp_source(
  tp_source      = tp_source,
  tunnel_lengths = c(th, r_len, r_len, th)
)

lobstr::tree(
  rapply(valid_tp, function(x) round(x, 3), how = "list")
)
```

The validator expands every state's inner list to the full canonical destination
set. Destinations supplied as `NULL` or omitted entirely are filled with zero
vectors of the appropriate length. For example, backwards slots not explicitly
provided — such as `stage_3$stage_1` (two-step reversal, not modelled here) and
`stage_2$stage_4` (skip-forward, not modelled) — are auto-added as zero vectors.

# Early resolution and progression within relapse tunnels

Because transition probabilities are non-zero at `d = 1` and `d = 2`, patients
can exit the relapse tunnel before reaching its final cycle. We can inspect the
full probability table at each `d`:

```{r stage2_exit}
# Probabilities out of stage_2 at each tunnel cycle d (rows = destinations):
s2_exit <- do.call(rbind, lapply(valid_tp$stage_2, identity))
rownames(s2_exit) <- names(valid_tp$stage_2)
round(t(s2_exit), 3)
```

```{r stage2_pstay}
# Implied p_stay at each cycle of stage_2:
1 - rowSums(round(t(s2_exit), 3))
```

```{r stage3_exit}
# Probabilities out of stage_3 at each tunnel cycle d:
s3_exit <- do.call(rbind, lapply(valid_tp$stage_3, identity))
rownames(s3_exit) <- names(valid_tp$stage_3)
round(t(s3_exit), 3)
```

```{r stage3_pstay}
# Implied p_stay at each cycle of stage_3:
1 - rowSums(round(t(s3_exit), 3))
```

At `d = 1` and `d = 2` the row sums are less than 1, confirming that most
patients remain in the relapse that cycle. The probabilities rise across `d`,
reflecting increasing clinical pressure to resolve or progress as the relapse
continues. At `d = r_len = 3`, the row sums are exactly 1 — the
decision-tree node forces every remaining patient out of the relapse.

```{r dtn_check}
# Confirm decision-tree node: row sums at d = r_len must equal 1
rowSums(round(t(s2_exit), 3))[r_len]
rowSums(round(t(s3_exit), 3))[r_len]
```

This design is richer than a pure decision-tree-with-delay (where all
probability mass sits only at the last cycle): genuine early exits occur at
`d = 1` and `d = 2`, but the DTN at `d = 3` guarantees a clean, complete exit
for any patients who have not yet resolved or progressed.

# Progressive phase is one-way

`stage_4` contains no backwards transitions. Patients who enter the secondary
progressive phase remain there — the only exit is death.

```{r stage4_tps}
round(valid_tp$stage_4$die, 3)
```

This is a deliberate modelling choice reflecting the current clinical
understanding that secondary progressive MS is irreversible. Note, however, that
this is a constraint imposed by the modeller through the transition probabilities
— not by `treatseqr` itself. A modeller working in a different disease area, or
using a different assumption, could supply non-zero backwards transitions for
`stage_4` without any changes to the model specification or the framework.

# Generating M and extrapolating

As with the other vignettes, use the function to generate our matrix list:

```{r generate_m}
m <- generate_m_list(
  m_specification      = m_spec,
  transition_prob_list = valid_tp,
  state_names          = c(
    "newly_dx", "stage_1", "stage_2", "stage_3", "stage_4", "die"
  )
)
```

Before extrapolating, it is instructive to visualise `m2` directly. In addition to
the superdiagonal strips (patients advancing along tunnels) and rightward
vertical strips (forward transitions to later tunnels), the relapse blocks
contain leftward vertical strips — the backwards transitions. These are absent
in the purely progressive models presented in the other vignettes.

```{r plot_m2}
m2_df <- as.data.frame(Matrix::summary(m$m2))

# label each entry by source tunnel block
s1_end <- 1L + th
s2_end <- s1_end + r_len
s3_end <- s2_end + r_len
s4_end <- s3_end + th

m2_df$source <- cut(
  m2_df$i,
  breaks = c(0L, s1_end, s2_end, s3_end, s4_end, s4_end + 1L),
  labels = c("stage_1", "stage_2", "stage_3", "stage_4", "death")
)

ggplot2::ggplot(
  m2_df,
  ggplot2::aes(x = j, y = i, colour = source, alpha = x)
) +
  ggplot2::geom_point(size = 2) +
  ggplot2::scale_y_reverse() +
  ggplot2::scale_alpha_continuous(range = c(0.3, 1), name = "Probability") +
  ggplot2::labs(
    title    = "Sparsity pattern of m2",
    subtitle = "Backwards strips appear to the left of each relapse block",
    x        = "Destination (column j)",
    y        = "Source (row i)",
    colour   = "Source state"
  ) +
  ggplot2::theme_minimal()
```

In the `stage_2` block (rows `r th + 2`–`r th + 4`) the leftward strip at column
2 represents resolution back to `stage_1`. In the `stage_3` block (rows
`r th + 5`–`r th + 7`) the leftward strip at column `r th + 2` represents partial
resolution back to `stage_2`. The `stage_1` block has no leftward strip — it is
the first tunnel so there is no earlier tunnel to resolve into. The `stage_4`
block similarly has no backwards strip, reflecting the one-way nature of
secondary progressive disease. The alpha scale (opacity) encodes probability:
within each relapse block the strips intensify toward `d = r_len` (the
decision-tree node), visible as lighter points at early d and darker points at
`d = 3`.

Finally, we can extrapolate the trace:

```{r extrapolate}
trace <- extrapolate_treatseqr(m = m, spec = m_spec)
```

A quick sanity check — column sums of the trace should equal 1 throughout:

```{r check_trace}
colSums(trace)
```

# Consolidation and outputs

Consolidate the expanded trace to something we're used to seeing in health
economics:

```{r consolidate}
res <- consolidate_treatseqr_trace(
  full_trace  = trace,
  spec        = m_spec,
  state_names = m$state_names
)
```

## State populations over time (`t`)

As with the other vignettes, we can plot the traces:

```{r plot_t}
graphics::matplot(
  res$t,
  type = "l", lty = 1,
  xlab = "Cycle", ylab = "Population",
  main = "State populations over absolute time"
)
graphics::legend(
  "topright",
  legend = colnames(res$t),
  col    = seq_len(ncol(res$t)),
  lty    = 1,
  cex    = 0.8
)
```

Several features of this trace are worth noting:

- **`newly_dx`** drains rapidly — almost all patients have entered `stage_1` by
  cycle 2.
- **`stage_1` remains populated** throughout the horizon. Because patients can
  resolve from `stage_2` back to `stage_1`, this state is continuously
  replenished and never fully drains, unlike in a purely progressive model.
- **`stage_2`** is also sustained longer than in a strictly forward model. The
  feedback loop — patients moving `stage_1 → stage_2 → stage_1 → stage_2 → …`
  — keeps replenishing the moderate-relapse population.
- **`stage_4`** grows gradually as severe relapses progress to secondary
  progressive disease; from there, the only exit is death.

Note: you can experiment with longer time horizons to see what happens over a
longer time period. To do this, just change `th` right at the start of this
vignette and run through. The trace looks a bit more like "the norm" when you
set it to `200`. The reason it is set to `r th` is so that the plot of `m2` is
clearer!

## Within-tunnel survival (`d`) and the feedback-loop problem

The `d` output captures the population distribution across `d = 1, …` for each
tunnel at each model cycle. In the absence of any re-entries into a tunnel, the
`rowSums` of a tunnel's rows in the full trace matrix would give a clean
within-tunnel survival curve.

In this model, however, patients can **re-enter** both `stage_1` and `stage_2`
via backwards transitions:

- `stage_2 → stage_1` feeds patients back to `d = 1` of `stage_1`, mixing
  re-entry patients with patients on their first visit.
- `stage_3 → stage_2` feeds patients back to `d = 1` of `stage_2`, similarly
  contaminating its `d = 1` slot.

As a result, `res$d` for `stage_1` and `stage_2` accumulates contributions from
multiple cohorts (first-entry and re-entries). However, unlike the case where
patients are held at the end of tunnels, the baseline value in `d = 0` for a
tunnel state is based on **total traffic** into the state. Consequently, the
`d`-based survival curves are actually valid **but** are polluted by people
entering the state at different points in wall time (`t`). Yet, because of the
fundamental "clock-reset" assumption in this package, `t` has no effect on
transitions in a tunnel state, meaning you *can* use these as within-state
survival curves. Just be aware that the disjoin between `t` and `d` may have
practical implications for mortality probabilities within the tunnels. The
implicit assumption is that the mortality only depends on `d`, and not `t`,
meaning that aging has no effect on them. This is another reason to jump to
patient-level simulation modelling: if the model is further complicated by
needing to capture some ever-growing `t`-based mortality rate, then the limit of
our approach is exceeded and patient simulation is a more appropriate option. In
the case whereby the mortality is likely to be higher in the tunnels than the
general population, then our approach still provides a substantial
simplification, at least from the perspective of the modeller (the statistics is
still complex).

```{r plot_d}
graphics::matplot(
  res$d,
  type = "l", lty = 1,
  xlab = "Cycles since tunnel entry", ylab = "Proportion remaining",
  main = "Within-tunnel d, case including backwards transitions"
)
graphics::legend(
  "topright",
  legend = colnames(res$d),
  col    = seq_len(ncol(res$d)),
  lty    = 1,
  cex    = 0.8
)
```


## Expected time in state

The total expected cycles spent in each state across the cohort:

```{r expected_time}
round(colSums(res$t), 2)
```
