---
title: "Sparse Matrix Extrapolation from First Principles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sparse Matrix Extrapolation from First Principles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(treatseqr)
```

This vignette demonstrates how to use sparse matrices to extrapolate "tunnel states" in a health economics context.
The transitions for the initial state (prior to the successive tunnels) change every cycle, and therefore the matrix multiplication is broken up into a component for the top row of the matrix `M` (`m1` for this top row), and another component for the successive tunnels (called `m2`).

## Parameters

We start by defining the time horizon (`th_cyc`) and the source transition probabilities. These probabilities would typically come from hazard extrapolations (e.g., from `flexsurv`).

```{r params}
th_cyc <- 50

# some simple time varying transition probabilities. in reality these will come
# from hazard extrapolations e.g. from `flexsurv`
tp_source <- list(
  on_1l = list(
    off_1l = seq(0.1, 0.05, length.out = th_cyc), # Example: linearly decreasing
    on_2l = seq(0.05, 0.1, length.out = th_cyc), # Example: linearly increasing
    off_2l = rep(0, th_cyc),
    bsc = rep(0, th_cyc),
    die = rep(0.01, th_cyc)
  ),
  off_1l = list(
    on_2l = seq(0.08, 0.02, length.out = th_cyc),
    off_2l = seq(0.01, 0.03, length.out = th_cyc),
    bsc = rep(0.02, th_cyc),
    die = seq(0.05, 0.1, length.out = th_cyc)
  ),
  on_2l = list(
    off_2l = seq(0.05, 0.02, length.out = th_cyc),
    bsc = rep(0.01, th_cyc),
    die = seq(0.1, 0.15, length.out = th_cyc)
  ),
  off_2l = list(
    bsc = seq(0.075, 0.05, length.out = th_cyc),
    die = seq(0.15, 0.2, length.out = th_cyc)
  ),
  bsc = list(
    die = seq(0.2, 0.3, length.out = th_cyc)
  )
)


# simulate what the list would look like for a time-varying TP:
tp_list <- lapply(tp_source, function(treatment_line) {
  p_stay <- 1 - Reduce(`+`, treatment_line)
  c(list(p_stay = p_stay), treatment_line)
})

# validate (all should be 1):
lapply(tp_list, function(treatment_line) {
  Reduce(`+`, treatment_line)
})
```

## Matrix Construction

We construct two matrices:
1. `m1`: Represents the transitions from the initial state to all possible destinations. This changes every model cycle.
2. `m2`: Represents the transitions for the successive tunnels.

```{r matrix_construction}
# Make an example m1 and m2.
# NOTE: m1 is the transition from initial state to all possible destinations.
# In this case patients can go from on_1l to any other state. nevertheless, this
# all populates the first row of the matrix:
matrix_size <- (th_cyc * (length(tp_list) - 1)) + 2 # +2 for bsc and death
j_m1 <- c(
  1, # Stay on_1l
  2, # go to off_1l
  th_cyc + 1, # go to on_2l
  (2 * th_cyc) + 1, # go to off_2l
  (3 * th_cyc) + 1, # go to bsc
  matrix_size # die
)
i_m1 <- rep(1, length(j_m1))

# so, m1 changes every model cycle - this makes a list of them
x_m1 <- lapply(1:th_cyc, function(cyc) {
  unlist(lapply(tp_list$on_1l, function(v) v[cyc]))
})


# let's stick to cycle 1:

model_cycle <- 1

i_m2 <- list(
  off_1l_stay = 2:(th_cyc + 1),
  off_1l_on_2l = 2:(th_cyc + 1),
  off_1l_off_2l = 2:(th_cyc + 1),
  off_1l_bsc = 2:(th_cyc + 1),
  off_1l_dead = 2:(th_cyc + 1),
  on_2l_stay = (th_cyc + 2):((2 * th_cyc) + 1),
  on_2l_off_2l = (th_cyc + 2):((2 * th_cyc) + 1),
  on_2l_bsc = (th_cyc + 2):((2 * th_cyc) + 1),
  on_2l_dead = (th_cyc + 2):((2 * th_cyc) + 1),
  off_2l_stay = ((2 * th_cyc) + 2):((3 * th_cyc) + 1),
  off_2l_bsc = ((2 * th_cyc) + 2):((3 * th_cyc) + 1),
  off_2l_dead = ((2 * th_cyc) + 2):((3 * th_cyc) + 1),
  bsc_stay = ((3 * th_cyc) + 2):((4 * th_cyc) + 1),
  bsc_dead = ((3 * th_cyc) + 2):((4 * th_cyc) + 1)
)

# the column (j) index is a function of the i index the time horizon and the nth
# tunnel. note that "stay" transitions are superdiagonal (1), whilst the
# transitions to move to other tunnels or death are vertical
j_m2 <- list(
  off_1l_stay = i_m2$off_1l_stay + 1,
  off_1l_on_2l = rep(th_cyc + 2, th_cyc),
  off_1l_off_2l = rep((2 * th_cyc) + 2, th_cyc),
  off_1l_bsc = rep((3 * th_cyc) + 2, th_cyc),
  off_1l_dead = rep(matrix_size, th_cyc),
  on_2l_stay = i_m2$on_2l_stay + 1,
  on_2l_off_2l = rep((2 * th_cyc) + 2, th_cyc),
  on_2l_bsc = rep((3 * th_cyc) + 2, th_cyc),
  on_2l_dead = rep(matrix_size, th_cyc),
  off_2l_stay = i_m2$off_2l_stay + 1,
  off_2l_bsc = rep((3 * th_cyc) + 2, th_cyc),
  off_2l_dead = rep(matrix_size, th_cyc),
  bsc_stay = i_m2$bsc_stay + 1,
  bsc_dead = rep(matrix_size, th_cyc)
)

x_m2 <- tp_list[c(
  "off_1l",
  "on_2l",
  "off_2l",
  "bsc"
)]

# use coordinates to define the matrices
m1 <- lapply(1:th_cyc, function(model_cycle) {
  Matrix::sparseMatrix(
    i = i_m1,
    j = j_m1,
    x = x_m1[[model_cycle]],
    dims = c(1, matrix_size)
  )
})

# now for the tunnel matrix:

m2 <- Matrix::sparseMatrix(
  i = unlist(i_m2, use.names = FALSE),
  j = unlist(j_m2, use.names = FALSE),
  x = unlist(x_m2, use.names = FALSE),
  dims = c(matrix_size, matrix_size)
)
m2[matrix_size, matrix_size] <- 1 # absorbing death state
```

## Visualization of Sparsity Pattern

We can visualize the sparsity pattern of the tunnel matrix `m2`.

```{r plot_sparsity, fig.width=7, fig.height=6}
# for illustration, let's make a plot of m2 to show the pattern:
transition_df <- data.frame(
  i = unlist(i_m2, use.names = FALSE),
  j = unlist(j_m2, use.names = FALSE),
  tunnel = rep(names(i_m2), times = sapply(i_m2, length))
)

# Merge with sum_m2 to assign tunnel labels only to nonzero entries
sum_m2 <- Matrix::summary(m2)
sum_m2 <- merge(sum_m2, transition_df, by = c("i", "j"), all.x = TRUE)

# Plot
ggplot2::ggplot(
  sum_m2,
  ggplot2::aes(x = j, y = i, color = tunnel, alpha = x)
) +
  ggplot2::geom_point() +
  ggplot2::scale_y_reverse() +
  ggplot2::labs(
    title = "Sparsity Pattern by Tunnel",
    x = "Column",
    y = "Row",
    color = "Tunnel"
  ) +
  ggplot2::theme_minimal()
```

## Simulation

Now we perform the simulation by iterating through the cycles.

```{r simulation}
pop <- matrix(0, nrow = matrix_size, ncol = th_cyc)
pop[1, 1] <- 1
for (cyc in 2:th_cyc) {
  pop[, cyc] <- as.numeric(pop[1, cyc - 1] %*% m1[[cyc]]) +
    as.numeric(pop[, cyc - 1] %*% m2)
}

# some assertions for validation:
assertthat::assert_that(
  all(abs(colSums(pop) - 1) < 1e-10),
  msg = "All time points should have a population that sums to 1"
)
```

## Results Analysis

We can aggregate the tunnel populations to produce a Markov trace and visualize the results.

```{r results}
# Summing up tunnel populations to produce a trace:
markov_trace <- data.table::data.table(
  cycle = 1:th_cyc,
  on_1l = pop[1, ],
  off_1l = colSums(pop[2:(th_cyc + 1), ]),
  on_2l = colSums(pop[(th_cyc + 2):((2 * th_cyc) + 1), ]),
  off_2l = colSums(pop[((2 * th_cyc) + 2):((3 * th_cyc) + 1), ]),
  bsc = colSums(pop[((3 * th_cyc) + 2):((4 * th_cyc) + 1), ]),
  dead = pop[matrix_size, ]
)

# all sum to 1 QC:
assertthat::assert_that(
  all(abs(rowSums(markov_trace[, -"cycle"]) - 1) < 1e-10),
  msg = "All time points should have a population that sums to 1"
)

# make plot data:
p_mk <- data.table::melt.data.table(
  markov_trace,
  id.vars = "cycle",
  value.vars = c("on_1l", "off_1l", "on_2l", "off_2l", "bsc", "dead")
)

# plot the trace:
plot_trace <- ggplot2::ggplot() +
  ggplot2::geom_line(
    data = p_mk,
    ggplot2::aes(
      x = cycle,
      y = value,
      color = variable
    ),
    linewidth = 1
  ) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = "Markov trace with treatment line tunnels",
    x = "Model cycle",
    y = "Proportion of cohort",
    color = "Health state"
  ) +
  ggplot2::scale_x_continuous(
    expand = ggplot2::expansion(mult = c(0, 0.05))
  ) +
  ggplot2::scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = ggplot2::expansion(mult = c(0, 0.05))
  )
```

Finally, we visualize the state residency over time for each tunnel state alongside the Markov trace.

```{r combined_plot, fig.width=10, fig.height=8}
# state residency plots by d (time within tunnel state) for each tunnel state

matplot_gg <- function(
  mat,
  start_row,
  end_row,
  xlab = "Time within tunnel (d)",
  ylim = NULL,
  ylab = "State residency",
  title = "Tunnel state residency over time",
  color = "black",
  ...
) {
  # mat: matrix with columns as series
  # color: optional vector of colors

  m <- t(mat[start_row:end_row, ])

  df <- data.table::data.table(m)
  df$t <- seq_len(nrow(df))
  df_long <- data.table::melt.data.table(
    df,
    id.vars = "t",
    variable.name = "series",
    value.name = "y"
  )

  if (is.null(ylim)) {
    ylim <- max(df_long$y)
  }

  ggplot2::ggplot() +
    ggplot2::geom_line(
      data = df_long,
      ggplot2::aes(x = t, y = y, color = series, group = series),
      colour = color
    ) +
    ggplot2::labs(x = xlab, y = ylab, title = title, color = "Series") +
    ggplot2::theme_minimal() +
    ggplot2::scale_x_continuous(
      expand = ggplot2::expansion(mult = c(0, 0.05))
    ) +
    ggplot2::scale_y_continuous(
      limits = c(0, ylim),
      labels = scales::percent_format(accuracy = 1),
      expand = ggplot2::expansion(mult = c(0, 0.05))
    )
}

library(patchwork)
ylim <- max(pop[2:((4 * th_cyc) + 1), ])
combined_plot <- plot_trace /
  (matplot_gg(
    mat = pop,
    start_row = 2,
    end_row = th_cyc + 1,
    ylim = ylim,
    title = "Off 1L tunnel state residency over time"
  ) |
    matplot_gg(
      mat = pop,
      start_row = th_cyc + 2,
      end_row = (th_cyc * 2) + 1,
      ylim = ylim,
      title = "Off 1L tunnel state residency over time"
    ) |
    matplot_gg(
      mat = pop,
      start_row = th_cyc + 3,
      end_row = (th_cyc * 3) + 1,
      ylim = ylim,
      title = "Off 1L tunnel state residency over time"
    ) |
    matplot_gg(
      mat = pop,
      start_row = th_cyc + 4,
      end_row = (th_cyc * 4) + 1,
      ylim = ylim,
      title = "Off 1L tunnel state residency over time"
    ))

combined_plot
```
