---
title: "Sparse Matrix Extrapolation using `treatseqr`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sparse Matrix Extrapolation using treatseqr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(treatseqr)
```

# Introduction

This vignette runs through the process of setting up and extrapolating a
sequencing model allowing patients to "skip over" tunnels (e.g. transition from
the first tunnel directly to the 3rd).  

This is accomplished by creating a sparse transition probability matrix (TPM),
populated with transitions in square "blocks" of size equal to the time horizon
(`th`), in addition to an extra row and column (or extra rows and columns) on
the top left for the initial health state(s) prior to the sequence of tunnels.
Visually, this looks like the below:

<img width="653" height="620" alt="image" src="https://github.com/user-attachments/assets/e32b19c8-e425-4a6a-9908-a03b1df4f1c1" />

Each "diagonal" is actually one to the right of the diagonal position inside
its block (termed superdiagonal (1)). This means that the probability of
staying in the "same state", that is the same time within the same tunnel, is
zero, and the probability of remaining in the tunnel is used to move the
patient "along" the tunnel by 1. For instance in the first tunnel if there is
one pre-tunnel state associated with column 1 and row 1, then cell 2,2 in the
matrix would be the probability of patients in the 1st cycle of the first
tunnel staying in the first cycle of the first tunnel, which is zero. INSTEAD,
the probability of remaining in the first tunnel after one cycle inside is
positioned in 2,3, which is the probability of moving from cycle 1 to cycle 2
of the first tunnel.  

Horizontally to the right of the superdiagonal elements are a series of
vertically arranged transition probabilities. These are the probabilities of a
patient transitioning to the _beginning_ of a subsequent tunnel. For instance,
in the first tunnel, the probability of moving from cycle 1 of tunnel 1 to
cycle 1 of tunnel 3 would be positioned in row 2, but in the same column as the
first row of the 3rd tunnel. During matrix multiplication, those patients then
"move" to the beginning of tunnel 3.  

If you look closely at the image, you can see that the first row is empty. This
is because the transition probabilities for the state(s) prior to entering the
sequence of tunnels are time-varying. Therefore, they need to be changed every
model cycle. It is computationally inefficient to do this to a large matrix
every cycle in `R`, due to its internal workings (it copies the whole matrix
upon modification). Therefore, we create a small matrix for the initial
state(s), comprising of one (or a few) rows and as many columns as required to
cover pre-states, all tunnels and the dead state. This smaller matrix is
referred to as `m1` in this package, and the associated function generates a
list of them to be used by the engine when extrapolating the sequencing model.  

The remaining set of transitions are for the tunnels and **do not change over
time**. This is because all within-tunnel time points, `d`, are actually
contained within one large sparse matrix (see the above image), so this only
needs to be compiled once. As this contains the empty row(s) associated with
the pre-tunnel sequence state(s), it is square. This matrix is referred to as
`m2` in this package, and the associated function generates this matrix to be
used by the engine when extrapolating the sequencing model.  

When considered together, the two matrices `m1` and `m2` are referred to as
`M`. A vector of population state residency `p` in `t-1` can be
matrix-multiplied by `M` for each model cycle. However, when `M` is broken down
into `m1` and `m2`, the operation is still relatively simple. `p1 = p *%* M` is
actually equivalent to `p1 = (p[1] %*% m1) + (p %*% m2)`. Similarly, this idea
can be further broken down on the tunnel level to allow for more efficient
computation. In the one pre-tunnel case for `k` tunnels, appropriate indices
`s`, and time horizon `th`, `p1 = (p[1] %*% m1) + (p[s1] %*% mt1) + ... +
(p[sk] %*% mtk)`, providing a way to "break down" the matrix multiplication
into smaller multiplications.  

Visually, this is simply horizontally slicing up `m2` in the obvious places, so
there is an `mt` which applies to patients inside of each tunnel. Towards the
end, this vignette explores the computational advantages of doing this vs
having one `m2` which is defined using sparse matrix functions.

# Specifying `M`

The size of `M` is determined by several things:

- The number of pre-tunnel states, `n_pre`
- The death state, which is always 1
- The time horizon, `th`
- The number of tunnels, `k`

`M` is square, and can be defined using the function `specify_m`:

```{r specify_m}
m_spec <- treatseqr::specify_m(
  n_cycles = 5,
  n_tunnels = 3,
  pre_tunnel_states = 2
)
```

Taking a look at the results, we can see that the matrix $\mathbf{M}$ will be `r
m_spec$matrix_size`x`r m_spec$matrix_size` in dimension. The function also
provides the coordinates within matrices `m1` and `m2` for all the data points
that could be needed to populate them, even those that may be zero when it comes
time to place data in the right places. This allows an efficiency step whereby
the final coordinate matrix can filter out data points with values equal to
zero, resulting in increases to the sparsity of the matrix, reducing
computation.  

But before we get into all that, let's look at what the coordinate matrix looks
like for `m1`:

```{r print_spec}
print(m_spec$m1_ijx)
```

in the above, `i` is row, `j` is column, and `x` is value, which at this stage
we do not know yet. given that we set `pre_tunnel_states` to `2`, we need to
populate the top two rows of `M` every model cycle, as these change over time,
and the transitions are based on time from baseline `t` rather than time within
tunnel, `d`. By the time we generate a Markov trace, we will have generated a
`list` object with length equal to the number of model cycles, with one matrix
for each cycle. In this sense, `m1` takes a familiar approach to time-varying
Markov modelling in `R` by simply applying the next appropriate matrix to the
relevant population.  

`m2` is larger and a bit more involved. To explore it, let's take a look at the
top 20 rows:

```{r m2_example_first_20}
print(m_spec$m2_ijx[1:20, ])
```

There are different parts to this matrix, which correspond to different
"blocks". As we set `pre_tunnel_states` to `2` in this case, the top left cell
of the first block is row `3`, column `3`. This block corresponds to the first
tunnel, and the first superdiagonal (1) elements are in rows `3` to `7`, columns
`4` to `8`. These represent the probabilities of remaining in the first tunnel,
given time in the tunnel `d`.  

The probability of staying at the same time within tunnel (`d`) is always zero,
so the column index `j` is always one higher than the row index `i` for this
part:

```{r m2_example_top_5}
print(m_spec$m2_ijx[1:5, ])
```

For transitions to subsequent tunnels, these are arranged vertically to
the right.

The transition from the first tunnel to the second:

```{r m2_example_tunnel1_to_2}
print(m_spec$m2_ijx[6:10, ])
```

First tunnel to third tunnel:

```{r m2_example_tunnel1_to_3}
print(m_spec$m2_ijx[11:15, ])
```

and so on, until we finally get to transitions from the first tunnel to death:

```{r m2_example_tunnel1_to_death}
print(m_spec$m2_ijx[16:20, ])
```

we know it is death because `m_spec$matrix_size` is `r m_spec$matrix_size`, and
that matches the number in the `j` column of the matrix.  

The next thing to note is that in each tunnel block after the first, the row
indices increase by `n_cycles`, as do the column indices, but the number of
"destination" states reduces by one. This is because the implicit assumption in
this kind of model is that patients cannot return to a previous tunnel. However,
it is possible to incorporate backwards transitions if desired by adding in
**horizontally** aligned transition probabilities. this is not currently within
this package, but it would be possible to add it. Such transitions would
essentially meet the vertically arranged transitions in `m2` currently, and
would therefore move patients back to the start of a previous tunnel.  

The number of transition probabilities needed for tunnel 2 in this case is `15`,
as opposed to `20` for tunnel 1:

```{r m2_example_tunnel2}
print(m_spec$m2_ijx[21:35, ])
```

Similarly the third and final tunnel has less rows again:

```{r m2_example_tunnel3}
print(m_spec$m2_ijx[36:45, ])
```

Finally, `m2` needs an absorbing death state, which the function helps you
identify by putting `1` in the x column as it is known:

```{r m2_example_death}
print(m_spec$m2_ijx[46, ])
```



The next step is to populate `m1_ijx` with the data (i.e., `x` values), as well
as `m2_ijx`. A function `populate_m` is used to organise the transition
probability data into the right order, and then put that in the `x` column of
`m1_ijx` and `m2_ijx`.  

## Validating the transition probability data

The transition probability data should be validated first. This is what
the function `validate_tp_source` is for. The transition probabilities should be
organised in quite a sepcific way to work with this package, which the validator
function will help you to arrive at. If we use the same example dataset we used
for the vignette `sparse-matrix-extrapolation`, we can validate it and then
explore what we need to do next:


```{r transition_probabilities}
th_cyc <- 50

# some simple time varying transition probabilities. in reality these will come
# from hazard extrapolations e.g. from `flexsurv`
tp_source <- list(
  on_1l = list(
    off_1l = seq(0.1, 0.05, length.out = th_cyc), # Example: linearly decreasing
    on_2l = seq(0.05, 0.1, length.out = th_cyc), # Example: linearly increasing
    off_2l = rep(0, th_cyc),
    bsc = rep(0, th_cyc),
    die = rep(0.01, th_cyc)
  ),
  off_1l = list(
    on_2l = seq(0.08, 0.02, length.out = th_cyc),
    off_2l = seq(0.01, 0.03, length.out = th_cyc),
    bsc = rep(0.02, th_cyc),
    die = seq(0.05, 0.1, length.out = th_cyc)
  ),
  on_2l = list(
    off_2l = seq(0.05, 0.02, length.out = th_cyc),
    bsc = rep(0.01, th_cyc),
    die = seq(0.1, 0.15, length.out = th_cyc)
  ),
  off_2l = list(
    bsc = seq(0.075, 0.05, length.out = th_cyc),
    die = seq(0.15, 0.2, length.out = th_cyc)
  ),
  bsc = list(
    die = seq(0.2, 0.3, length.out = th_cyc)
  )
)
```


Looking at the `tp` values in a visual way, we can see this:

```{r view_data_tree}
lobstr::tree(
  rapply(tp_source, function(x) round(x, 2), how = "list")
)
```


This has some important features:

1. The first level elements are health states
2. The first level elements are in order, going from first to last tunnel
3. Each first level element contains transition probabilities to all other
   subsequent tunnels, even if zero probability (these can be NULL but must
exist to use the validator function). This means length goes down by 1 for each
successive tunnel
4. element wise sums of everything in each 1st level element (i.e., each health
   state) are all $\le 1$
5. the names of the elements all correspond to each other appropriately, no
   differences or spelling mistakes are allowed!

When we validate against these rules, the function returns us an object that can
be used to populate `m1` and `m2` (in this case the same as the data going in):

```{r validate_transition_probabilities}
valid_tp <- validate_tp_source(tp_source = tp_source)
lobstr::tree(
  rapply(valid_tp, function(x) round(x, 2), how = "list")
)
```


If instead, we put `NULL` for the empty transitions, the validator fills these
in:

```{r transition_probabilities_with_nulls}
th_cyc <- 50

# some simple time varying transition probabilities. in reality these will come
# from hazard extrapolations e.g. from `flexsurv`
tp_source <- list(
  on_1l = list(
    off_1l = seq(0.1, 0.05, length.out = th_cyc), # Example: linearly decreasing
    on_2l = seq(0.05, 0.1, length.out = th_cyc), # Example: linearly increasing
    off_2l = NULL,
    bsc = NULL,
    die = rep(0.01, th_cyc)
  ),
  off_1l = list(
    on_2l = seq(0.08, 0.02, length.out = th_cyc),
    off_2l = seq(0.01, 0.03, length.out = th_cyc),
    bsc = rep(0.02, th_cyc),
    die = seq(0.05, 0.1, length.out = th_cyc)
  ),
  on_2l = list(
    off_2l = seq(0.05, 0.02, length.out = th_cyc),
    bsc = rep(0.01, th_cyc),
    die = seq(0.1, 0.15, length.out = th_cyc)
  ),
  off_2l = list(
    bsc = seq(0.075, 0.05, length.out = th_cyc),
    die = seq(0.15, 0.2, length.out = th_cyc)
  ),
  bsc = list(
    die = seq(0.2, 0.3, length.out = th_cyc)
  )
)
lobstr::tree(
  rapply(tp_source, function(x) round(x, 2), how = "list")
)
```


compared to:

```{r validate_transition_probabilities_with_nulls}
# Fills in the blanks if they are NULLs
valid_tp <- validate_tp_source(tp_source = tp_source)
lobstr::tree(
  rapply(valid_tp, function(x) round(x, 2), how = "list")
)
```


However, the function will give you an error if you miss the elements out
completely. This prevents any kind of corner-cutting when specifying the model
and ensures that you are very carefully putting things together.


## A note on order of operations

The last step before entering the data into `M` is to make sure that
`valid_tp`, which is valid in and of itself, corresponds to the specification
of `M`.  

Well, if we reverse the order of operations, we can make the process automatic
by using `valid_tp` to tell `specify_m` the inputs it needs. These are:

- `n_cycles`
- `n_tunnels`
- `pre_tunnel_states`

`n_cycles` can be derived from `valid_tp` because all 2nd level elements have
the same length. However, the other two inputs `n_tunnels` and
`pre_tunnel_states` must be determined by the user, as the structure of the data
is identical whether or not a state is a pre-tunnel. There are the same amount
of destination states, and the `tp` vectors have the same lengths, names and so
on.   

Yet, it is still most robust to derive the validated `tp` data first, and then
specify `M`, because during that, some additional validation of `n_tunnels` and
`pre_tunnel_states` can be conducted. For instance, in this model framework,
there must be at least one tunnel, and one pre-tunnel state.

In most oncology settings, only the first state can be a pre-tunnel state
because the transition from e.g., off-treatment in first line to any subsequent
state has an index date of discontinuation event for first line therapy, not
baseline. However, there are feasibly use-cases for multiple pre-tunnel states.
Yet for this example, **we say that only `on_1l` is a pre-tunnel state**.

```{r specify_m_from_valid_tp}
m_spec <- specify_m(
  n_cycles = 50,
  n_tunnels = length(valid_tp) - 1,
  pre_tunnel_states = 1
)
```

## Populating `m1`

The baseline-time transitions will be a list of `r th_cyc` `m1`s in this case, one for
each cycle. Each of these is a matrix with `1` row and `r m_spec$matrix_size`
columns.

We can compile this manually if we want, by doing the following. First, pull out
the probabilities for the health state(s) we need to populate once per cycle:

```{r derive_m1_data}
p_off_1l <- valid_tp$on_1l$off_1l
p_on_2l <- valid_tp$on_1l$on_2l
p_off_2l <- valid_tp$on_1l$off_2l
p_bsc <- valid_tp$on_1l$bsc
p_die <- valid_tp$on_1l$die
m_size <- m_spec$matrix_size
m1_coord <- m_spec$m1_ijx
```

Next, have a look at the structure of `m1_ijx` to see where these need to go:

```{r print_m1_ijx}
print(m_spec$m1_ijx)
```

In this case:

- `1` is staying in `on_1l`
- `2` is discontinuation to `off_1l`
- `52` is starting `on_2l`
- `102` is starting `off_2l`
- `152` is starting `bsc`
- `203` is death

Therefore we can easily make our list like this:

```{r manual_m1_list}
m1_list <- lapply(seq_len(th_cyc), function(model_cycle) {
  m1_coord[, "x"] <- c(
    1 - sum(
      p_off_1l[model_cycle],
      p_on_2l[model_cycle],
      p_off_2l[model_cycle],
      p_bsc[model_cycle],
      p_die[model_cycle]
    ),
    p_off_1l[model_cycle],
    p_on_2l[model_cycle],
    p_off_2l[model_cycle],
    p_bsc[model_cycle],
    p_die[model_cycle]
  )
  m1_coord <- m1_coord[m1_coord[, "x"] > 0, , drop = FALSE]
  Matrix::sparseMatrix(
    i = m1_coord[, "i"],
    j = m1_coord[, "j"],
    x = m1_coord[, "x"],
    dims = c(1, m_size)
  )
})
print(m1_list[[1]])
```

This defines a sparse matrix for each model cycle describing the probabilities
for transition for that model cycle from the pre-tunnel state(s) to all other
states in the model. Note the step to remove 0 probabilities before creating the
sparse matrix. This then removes unnecessary data points and therefore
computation later on. To check all sum to 1 (which they must given the 1 - sum()
in the above!):

```{r assert_m1_sums}
assertthat::assert_that(
  all(unlist(lapply(m1_list, sum)) == 1),
  msg = "m1 rows do not sum to 1"
)
```

In practice, validation steps like these are not necessary because the TPs come
from `valid_tps`, which have already been validated against conditions like
non-negativity and summing to 1. The above is just to prove that it is working
correctly.

To make this a bit less cumbersome, we can use the function `generate_m_list`
to do the above using the validated `tp` list for `m1` and `m2` all in one easy
step. The result will be a list with 2 elements, one for `m1` one for `m2`, with
`m1` being a list of length time horizon, and `m2` being a single sparse matrix.

```{r generate_m_list}
# validate inputs and generate the full matrix m from one function
m <- generate_m_list(
  m_specification = m_spec,
  transition_prob_list = tp_source
)
```

Note that this function automatically filters out probabilities entered as zero.
This then avoids unnecessary computation when running extrapolations, which
become more important in e.g., probabilistic settings. When running a
probabilistic sensitivity analysis, one would need to generate `tp_source` `n`
times, where `n` is the number of iterations, and then generate `m` for each
iteration. Therefore, the efficiency of this step is important to consider.


# Extrapolation

From this point it is very simple to generate a Markov trace!

```{r extrapolate_manual}
# output population matrix:
pop <- matrix(0, nrow = m_spec$matrix_size, ncol = th_cyc + 1)
pop[1, 1] <- 1

# get the TPs
m1 <- m$m1
m2 <- m$m2

# extrapolate the model:
for (cyc in seq_len(th_cyc) + 1) {
  pop[, cyc] <-
    as.numeric(pop[seq_len(m_spec$pre_tunnels), cyc - 1] %*% m1[[cyc - 1]]) +
    as.numeric(pop[, cyc - 1] %*% m2)
}
```

That's it, for the most part. The next step is consolidating information from
this matrix `pop`, which has some useful properties.

## Properties of `pop`

`pop` is an interesting object because you can add up bits of it to get
different bits of information. These are really useful for cost-effectiveness
modelling.

### Overall survival

One minus the last row is overall survival:

```{r plot_os}
plot(1 - pop[nrow(pop), ],
  type = "l", xlab = "Cycle", ylab = "Overall Survival",
  main = "Overall Survival Curve"
)
```

### Tunnel populations in absolute time `t`

Vertical sums of sections of `pop` each of height of time horizon, except for
the pre-tunnel states and death (one row each), are tunnel state populations at
each absolute time point, `t`, so in cycle 2, we have `colSums(pop[2:51, ])[2]`,
which is `r colSums(pop[2:51, ])[2]` i.e., `1` (baseline pop) multiplied by the
probability of discontinuing first-line but not starting second line (`0.1`).  

In English, 0.1 of the cohort is in the first tunnel state in cycle 2.  

Plotting this is easy if you know which rows to add up. In the below we have
entered the row numbers manually to help you see the blocks for yourselves. Note
that the time horizon is 50 cycles in this example:

```{r plot_tunnel_pops}
plot(1 - pop[202, ],
  type = "l",
  col = "black",
  xlab = "Cycle", ylab = "Tunnel 1 Population",
  main = "Tunnel Populations over Time"
)
lines(colSums(pop[2:51, ]), type = "l", col = "grey80")
lines(colSums(pop[52:101, ]), type = "l", col = "red")
lines(colSums(pop[102:151, ]), type = "l", col = "blue")
lines(colSums(pop[152:201, ]), type = "l", col = "green")
```


### Tunnel populations in relative time `d`

This is where `pop` gets more interesting. The sum _across_ rows is the
population from the time point of entering the tunnel, or in other words `d = 0`
for that tunnel. For instance, in the above example, the sum of the 2nd row is
the proportion of the cohort that moved into tunnel 1 at any time in the model.  

If you take the appropriate set of `rowSum` results and divide it by the
first element of the result, this will give you a within-tunnel survival (i.e.,
population starting from 100% remaining inside of the tunnel in question).  

This lets you plot something more familiar, but only if you divide the values by
their first element. Without that, it looks strange:

```{r plot_d_unadjusted}
plot(1 - pop[202, ],
  type = "l",
  col = "black",
  xlab = "Cycle", ylab = "Tunnel 1 Population",
  main = "Tunnel Populations over Time"
)
lines(rowSums(pop[2:51, ]), type = "l", col = "grey80")
lines(rowSums(pop[52:101, ]), type = "l", col = "red")
lines(rowSums(pop[102:151, ]), type = "l", col = "blue")
lines(rowSums(pop[152:201, ]), type = "l", col = "green")
```

But if you adjust it for the proportion of the initial cohort that ever entered
each tunnel, you get something more familiar:

```{r plot_d_adjusted}
plot(1 - pop[202, ],
  type = "l",
  col = "black",
  xlab = "Cycle", ylab = "Tunnel 1 Population",
  main = "Tunnel Populations over Time"
)
lines(rowSums(pop[2:51, ]) / sum(pop[2, ]), type = "l", col = "grey80")
lines(rowSums(pop[52:101, ]) / sum(pop[52, ]), type = "l", col = "red")
lines(rowSums(pop[102:151, ]) / sum(pop[102, ]), type = "l", col = "blue")
lines(rowSums(pop[152:201, ]) / sum(pop[152, ]), type = "l", col = "green")
```

This is the **within-tunnel survival curve** for each tunnel, which is much more
intuitive, and is more likely to be comparable to some external data for
external validation purposes.  

However, one warning here is that the source Kaplan-Meier used to inform the
transition probabilities for these tunnel states **will not align** with these
curves. This is because the appropriate endpoint for informing the transition
probabilities needs to have carefully designed censoring rules to isolate the
hazard of a specific transition. Therefore, these curves, which also include
exiting the state for _any_ reason (not just one specific event) should be
compared to Kaplan-Meier which **does not e.g., censor for death**. In other
words, these extrapolations should be compared to within-state survival, not
e.g., time to progression. 

As long as you carefully think about what data these lines should be compared
to first, you can use these survival at time `d` for each tunnel plots to
internally or even externally validate your model's extrapolation of clinical
outcomes.


## Functions to generate outputs

The full matrix is the most useful object to return for `pop`. However,
interpreting it then requires manual steps to add things up in ways that
typically will always be needed for cost-effectiveness modelling in health
economics. Yet, individual preferences will always be a factor, so we have
broken down the steps into separate simple functions to do this for you.

### Extrapolation

The extrapolation function is trivial, but still useful to have. This function
extrapolates the model, returning the full trace for you to do with as you
please:

```{r extrapolate}
trace <- extrapolate_treatseqr(m = m, spec = m_spec)
colSums(trace)
```


### Consolidation

After the trace runs, we need to consolidate the results into useful outputs.
This usually includes both the vertical and horizontal sums (see above). The
function `consolidate_treatseqr_trace` does this for you, returning a list
containing both.

Interestingly, the bounds of the states within `trace` always correspond to the
numbers in `m1_ijx` in the `j` column, which simplifies the process of adding
things up nicely. We just use a simple algorithm leveraging information stored
in `m_spec` to figure out which pairings of values in `m1_ijx[, "j"]` correspond
to which state:

- `pre_tunnels`: The first `pre_tunnels` rows in `trace` are 1 row/column each
and these are states which have a basis in time of `t` not `d`, so are simply
the appropriate row (e.g., `trace[1, ]` is the first health state over time).
- Tunnels: Each tunnel has `n_cycles` rows/columns, and these can be summed to
produce what we need. Note that we can do horizontal and vertical sums here -
the bounds are just the transpose of each other.
- Death: The last row/column is death, which is absorbing.


```{r consolidate}
trace <- consolidate_treatseqr_trace(
  full_trace = trace,
  spec = m_spec,
  state_names = m$state_names
)

# expected time in state in model cycles:
colSums(trace$t)
```

These can be used for easy plotting, using `matplot` from `base`.

```{r matplot_t}
graphics::matplot(trace$t, type = "l")
graphics::legend(
  "topright",
  legend = colnames(trace$t),
  col = seq_len(ncol(trace$t)),
  lty = 1
)
```



Similarly, we can look at the proportional survival in time in state `d`.

```{r matplot_d}
graphics::matplot(trace$d, type = "l")
graphics::legend(
  "topright",
  legend = colnames(trace$d),
  col = seq_len(ncol(trace$d)),
  lty = 1
)
```
