---
title: "Sparse Matrix Extrapolation using `treatseqr`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sparse Matrix Extrapolation using treatseqr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(treatseqr)
```

# Introduction

This vignette runs through the process of setting up and extrapolating a
sequencing model allowing patients to "skip over" tunnels (e.g. transition from
the first tunnel directly to the 3rd), and also to "skip back" to any previous
tunnel (e.g., from the 2nd tunnel back to the first). For technical reasons
(wall time vs sojourn time), backwards transitions to pre-tunnel states are not
supported. To do this, it is more appropriate to use a microsimulation that
tracks how many times individual patients have attended a previous state, and
for how long. Without substantially complicating our approach, this can't be
done with simple matrices whilst balancing computational demands (to our
knowledge).

This is accomplished by creating a sparse transition probability matrix (TPM),
populated with transitions in square "blocks" of size equal to the time horizon
(`th`), in addition to an extra row and column (or extra rows and columns) on
the top left for the initial health state(s) prior to the sequence of tunnels.
Visually, this looks like the below:

<img width="653" height="620" alt="image" src="https://github.com/user-attachments/assets/e32b19c8-e425-4a6a-9908-a03b1df4f1c1" />

Each "diagonal" is actually one to the right of the diagonal position inside
its block (termed superdiagonal (1)). This means that the probability of
staying in the "same state", that is the same time within the same tunnel, is
zero, and the probability of remaining in the tunnel is used to move the
patient "along" the tunnel by 1. For instance in the first tunnel if there is
one pre-tunnel state associated with column 1 and row 1, then cell 2,2 in the
matrix would be the probability of patients in the 1st cycle of the first
tunnel staying in the first cycle of the first tunnel, which is zero. INSTEAD,
the probability of remaining in the first tunnel after one cycle inside is
positioned in 2,3, which is the probability of moving from cycle 1 to cycle 2
of the first tunnel.  

On either side of the superdiagonal elements are a series of vertically arranged
transition probabilities. These are the probabilities of a patient transitioning
to the beginning of any other tunnel — either a later tunnel (strip to the
right) or an earlier one (strip to the left, i.e. a backwards transition). For
instance, in the first tunnel, the probability of moving from cycle 1 of tunnel
1 to cycle 1 of tunnel 3 would be positioned in row 2, but in the same column as
the first row of the 3rd tunnel. During matrix multiplication, those patients
then "move" to the beginning of tunnel 3.  

If you look closely at the image, you can see that the first row is empty. This
is because the transition probabilities for the state(s) prior to entering the
sequence of tunnels are time-varying. Therefore, they need to be changed every
model cycle. It is computationally inefficient to do this to a large matrix
every cycle in `R`, due to its internal workings (it copies the whole matrix
upon modification). Therefore, we create a small matrix for the initial
state(s), comprising of one (or a few) rows and as many columns as required to
cover pre-states, all tunnels and the dead state. This smaller matrix is
referred to as `m1` in this package, and the associated function generates a
list of them to be used by the engine when extrapolating the sequencing model.  

The remaining set of transitions are for the tunnels and **do not change over
time**. This is because all within-tunnel time points, `d`, are actually
contained within one large sparse matrix (see the above image), so this only
needs to be compiled once. As this contains the empty row(s) associated with
the pre-tunnel sequence state(s), it is square. This matrix is referred to as
`m2` in this package, and the associated function generates this matrix to be
used by the engine when extrapolating the sequencing model.  

When considered together, the two matrices `m1` and `m2` are referred to as
`M`. A vector of population state residency `p` in `t-1` can be
matrix-multiplied by `M` for each model cycle. However, when `M` is broken down
into `m1` and `m2`, the operation is still relatively simple. `p1 = p *%* M` is
actually equivalent to `p1 = (p[1] %*% m1) + (p %*% m2)`. Similarly, this idea
can be further broken down on the tunnel level to allow for more efficient
computation. In the one pre-tunnel case for `k` tunnels, appropriate indices
`s`, and time horizon `th`, `p1 = (p[1] %*% m1) + (p[s1] %*% mt1) + ... +
(p[sk] %*% mtk)`, providing a way to "break down" the matrix multiplication
into smaller multiplications.  

Visually, this is simply horizontally slicing up `m2` in the obvious places, so
there is an `mt` which applies to patients inside of each tunnel. Towards the
end, this vignette explores the computational advantages of doing this vs
having one `m2` which is defined using sparse matrix functions.

# Specifying `M`

The size of `M` is determined by several things:

- The number of pre-tunnel states, `n_pre`
- The death state, which is always 1
- The time horizon, `th`
- The number of tunnels, `k`

`M` is square, and can be defined using the function `specify_m`:

```{r specify_m}
m_spec <- treatseqr::specify_m(
  tunnel_lengths = rep(5, 3),
  pre_tunnel_states = 2
)
```

Taking a look at the results, we can see that the matrix $\mathbf{M}$ will be `r
m_spec$matrix_size`x`r m_spec$matrix_size` in dimension. The function also
provides the coordinates within matrices `m1` and `m2` for all the data points
that could be needed to populate them, even those that may be zero when it comes
time to place data in the right places. This allows an efficiency step whereby
the final coordinate matrix can filter out data points with values equal to
zero, resulting in increases to the sparsity of the matrix, reducing
computation.  

But before we get into all that, let's look at what the coordinate matrix looks
like for `m1`:

```{r print_spec}
print(m_spec$m1_ijx)
```

Note that each pre-tunnel state has a set of coordinate slots allowing backwards
transitions to permit modellers to use this should they need it. The required
transitions are essentially from a pre-tunnel state to all other states. In the
function, backwards slots default to zero and are filtered out before putting
numbers into the large matrix objects or extrapolating the model.

In the above, `i` is row, `j` is column, and `x` is value, which at this stage
we do not know yet. Given that we set `pre_tunnel_states` to `2`, we need to
populate the top two rows of `M` every model cycle, as these change over time.
In other words, the pre-tunnel states are the states before the treatment
sequence, and their transition probabilities are based on time from baseline,
`t`, rather than time within each tunnel, `d` (where there is a separate `d` for
each tunnel, and `d` "resets" if patients loop back into a tunnel they were
previously resident within).

For `m1`, we will generate a `list` object with length equal to the number of
model cycles, with one matrix for each cycle. In this sense, `m1` takes a
familiar approach to time-varying Markov modelling in `R` by simply applying the
next appropriate matrix to the relevant population. This is no different from
defining a 3-dimensional array which would look like this: `m1[row, column,
cycle]`, but the `list` approach is more efficient in `R` because it avoids the
need to copy the whole array every time we change one value.

`m2` is larger and a bit more involved. To explore it, let's take a look at the
top 20 rows:

```{r m2_example_first_20}
print(m_spec$m2_ijx[1:20, ])
```

There are different "chunks" of information in `m2` corresponding to different
tunnels, usually called **blocks**. As we set `pre_tunnel_states` to `2` in this
case, the top left cell of the first block is cell `3,3` in the full matrix `M`.
This block corresponds to the first tunnel, and the first superdiagonal (1)
elements are in rows `3` to `7`, columns `4` to `8`. These represent the
probabilities of remaining in the first tunnel but moving "along" by 1 time
slot within the tunnel, given time in the tunnel `d`. By contrast, the diagonal
elements `[3,3]` to `[7,7]` represent the probability of staying within the
tunnel but also staying at the same `d`, essentially "freezing time" within the
tunnel. This is very important to understand when considering our approach to
fixed-duration tunnel-states, because `treatseqr` allows users to set a
probability of staying in the last `d` of the tunnel state. Usually, modellers
would set this probability to 0 by making the probability of going *somewhere*
equal to 1 at the end of a tunnel in something akin to a decision-tree node.
However, `treatseqr` is designed to avoid being overly prescriptive, and
therefore allows users to set a probability of remaining in the last `d` of the
tunnel. This gives a mechanism for "holding" patients within a tunnel until they
exit it if there is a health state which matches this conceptual design. Note
that this would make an exponential transition probability assumption for this
"tunnel-within_tunnel" state as the same transition probabilities would be
applied subsequently for these patients. For this reason, we would recommend
building in a decision-tree node at the end of short fixed-duration tunnel
states, such that the probability of staying in the tunnel past its end is zero.

Practically, this is simple to implement as one would simply put the
decision-tree node probabilities at the end of the vector for the tunnel (e.g.,
if the tunnel was 3 cycles long, `p_stay` could be `c(0.9,0.8,0)`, `p_next`
could be `c(0.05, 0.1, 0.9)` and `p_die` could be `c(0.05, 0.1, 0.1)` such that
they all add up to 1 and the probability of staying in the tunnel is zero).

The eagle-eyed among you may have seen that the function `specify_m` produces
some probabilities on the diagonals:

```{r m2_example_first_20}
print(m_spec$m2_ijx[which(m_spec$m2_ijx[, 1] == m_spec$m2_ijx[, 2]), ])
```

This corresponds to the above discussion (except for the last one, which is
death). Essentially, the last transition in each tunnel state **is applied to
the diagonal not superdiagonal**. This keeps patients within the tunnel if the
probability is `>0`, rather than making structural assertions on your
cost-effectiveness model, which may not follow the conceptual design forcing
patients from one tunnel into the next one in the sequence!

As a simple(ish) example, imagine a treatment sequence with 4 tunnels and one
pre-tunnel for newly diagnosed (i.e., 5 treatment lines). What if the 3rd line
therapy (tunnel 2) was a surgical intervention with a 3 cycle recovery period,
after which patients are then split into responders (tunnel 3) and
non-responders (tunnel 4)? Well, with `treatseqr` you can easily do this. With a
time horizon of 5 (to keep the print short), this would be:

```{r example_specify}
treatseqr::specify_m(
  tunnel_lengths = c(5, 5, 3, 5, 5),
  pre_tunnel_states = 1
)
```

With some simple treatment of the probabilities at the end of the 2nd tunnel
state, we can put a decision-tree node at the end of the tunnel that is 3 cycles
long such that patients either move to the 3rd or 4th tunnel.

Except for the tunnel ends, the diagonal elements are all zero, so are excluded
from the coordinates, and the values we need go into elements with `j = i + 1`:

```{r m2_example_top_5}
print(m_spec$m2_ijx[1:4, ])
```

For transitions to all other tunnels, values are arranged as vertical strips.
Strips to the right of the superdiagonal represent forward transitions; strips
to the left represent backwards transitions. Patients exiting the tunnel go to
the beginning of the destination tunnel regardless of direction.

The transition from the first tunnel to the second:

```{r m2_example_tunnel1_to_2}
print(m_spec$m2_ijx[6:10, ])
```

First tunnel jumping to third tunnel:

```{r m2_example_tunnel1_to_3}
print(m_spec$m2_ijx[11:15, ])
```


Modelling-wise this is incredibly useful, as **the model is not limited to a
series of successive tunnels**. Patients can skip to later tunnels. This is
highly relevant in oncology, where it is well known that patients opt out of
further treatment in favor of living life away from medical care as much as
possible, especially as prognoses become worse. This gives a simple mechanism
for this conceptual design. Simply supply transition probabilities from ANY
state to a "no treatment" tunnel (or several for different situations), and then
assign the correct probabilities in the right places. Then, `treatseqr` will
generate a Markov trace allowing for the often overlooked impact of later-line
opting out of treatment on overall cohort outcomes.


Finally, the death state transitions from the first tunnel to death in this case
are:

```{r m2_example_tunnel1_to_death}
print(m_spec$m2_ijx[16:20, ])
```

We know columns `r m_spec$matrix_size` is death because `m_spec$matrix_size` is
`r m_spec$matrix_size`, and that matches the number in the `j` column of the
matrix.  

Every tunnel block has the same number of vertical strips — one for each other
tunnel plus death, including backwards transitions, even if you haven't provided
any. For example, tunnel 2 can transition to the beginning of tunnel 1 just as
easily as forward to tunnel 3 or to death. Backwards destinations default to
zero if not supplied and are filtered from the sparse matrix before
construction, so there is very little if any computational cost (nanoseconds)
for unused transitions. This is particularly useful in e.g. relapsing-remitting
disease areas where a relapse tunnel can resolve without progression (backwards)
or with progression (forwards).

In `treatseqr`, we are agnostic to progressive disease structures, so the number
of transition probabilities needed for every tunnel is the same. In this case it is
`20`:

```{r m2_example_tunnel2}
print(m_spec$m2_ijx[21:40, ])
```

Similarly the third tunnel also has 20 rows, now including backwards strips to
both tunnel 1 and tunnel 2:

```{r m2_example_tunnel3}
print(m_spec$m2_ijx[41:60, ])
```

Finally, `m2` needs an absorbing death state, which the function helps you
identify by putting `1` in the x column as it is known:

```{r m2_example_death}
print(m_spec$m2_ijx[nrow(m_spec$m2_ijx), ])
```

The next step is to populate `m1_ijx` with the data (i.e., `x` values), as well
as `m2_ijx`. A function `populate_m` is used to organise the transition
probability data into the right order, and then put that in the `x` column of
`m1_ijx` and `m2_ijx`.  

## Validating the transition probability data

The transition probability data should be validated first. This is what the
function `validate_tp_source` is for. The transition probabilities should be
organised in quite a specific way to work with this package, which the validator
function will help you to arrive at. If we use the same example dataset we used
for the vignette `sparse-matrix-extrapolation`, we can validate it and then
explore what we need to do next:


```{r transition_probabilities}
th_cyc <- 50

# some simple time varying transition probabilities. in reality these will come
# from hazard extrapolations e.g. from `flexsurv`
tp_source <- list(
  on_1l = list(
    off_1l = seq(0.1, 0.05, length.out = th_cyc), # Example: linearly decreasing
    on_2l = seq(0.05, 0.1, length.out = th_cyc), # Example: linearly increasing
    off_2l = rep(0, th_cyc),
    bsc = rep(0, th_cyc),
    die = rep(0.01, th_cyc)
  ),
  off_1l = list(
    on_2l = seq(0.08, 0.02, length.out = th_cyc),
    off_2l = seq(0.01, 0.03, length.out = th_cyc),
    bsc = rep(0.02, th_cyc),
    die = seq(0.05, 0.1, length.out = th_cyc)
  ),
  on_2l = list(
    off_2l = seq(0.05, 0.02, length.out = th_cyc),
    bsc = rep(0.01, th_cyc),
    die = seq(0.1, 0.15, length.out = th_cyc)
  ),
  off_2l = list(
    bsc = seq(0.075, 0.05, length.out = th_cyc),
    die = seq(0.15, 0.2, length.out = th_cyc)
  ),
  bsc = list(
    die = seq(0.2, 0.3, length.out = th_cyc)
  )
)
```


Looking at the `tp` values in a visual way, we can see this:

```{r view_data_tree}
lobstr::tree(
  rapply(tp_source, function(x) round(x, 2), how = "list")
)
```


This has some important features:

1. The first level elements are health states
2. The first level elements are in order, going from first to last tunnel
3. Each first level element contains transition probabilities to all other
   tunnels and the pre-tunnels (these can be excluded if not needed)
4. Element wise sums of everything in each 1st level element (i.e., each health
   state) are all $\le 1$ such that the rows still sum to 1.
5. The names of the elements all correspond to each other appropriately, no
   differences or spelling mistakes are allowed!

When we validate against these rules, the function returns us an object that can
be used to populate `m1` and `m2`:

```{r validate_transition_probabilities}
valid_tp <- validate_tp_source(
  tp_source = tp_source,
  tunnel_lengths = rep(th_cyc, length(tp_source) - 1L),
  state_names = c(names(tp_source), "die")
)
lobstr::tree(
  rapply(valid_tp, function(x) round(x, 2), how = "list")
)
```


If instead, we put `NULL` for the empty transitions, the validator fills these
in:

```{r transition_probabilities_with_nulls}
th_cyc <- 50

# some simple time varying transition probabilities. in reality these will come
# from hazard extrapolations e.g. from `flexsurv`
tp_source <- list(
  on_1l = list(
    off_1l = seq(0.1, 0.05, length.out = th_cyc), # Example: linearly decreasing
    on_2l = seq(0.05, 0.1, length.out = th_cyc), # Example: linearly increasing
    off_2l = NULL,
    bsc = NULL,
    die = rep(0.01, th_cyc)
  ),
  off_1l = list(
    on_2l = seq(0.08, 0.02, length.out = th_cyc),
    off_2l = seq(0.01, 0.03, length.out = th_cyc),
    bsc = rep(0.02, th_cyc),
    die = seq(0.05, 0.1, length.out = th_cyc)
  ),
  on_2l = list(
    off_2l = seq(0.05, 0.02, length.out = th_cyc),
    bsc = rep(0.01, th_cyc),
    die = seq(0.1, 0.15, length.out = th_cyc)
  ),
  off_2l = list(
    bsc = seq(0.075, 0.05, length.out = th_cyc),
    die = seq(0.15, 0.2, length.out = th_cyc)
  ),
  bsc = list(
    die = seq(0.2, 0.3, length.out = th_cyc)
  )
)
lobstr::tree(
  rapply(tp_source, function(x) round(x, 2), how = "list")
)
```


compared to:

```{r validate_transition_probabilities_with_nulls}
# Fills in the blanks if they are NULLs
valid_tp <- validate_tp_source(
  tp_source = tp_source,
  tunnel_lengths = rep(th_cyc, length(tp_source) - 1L),
  state_names = c(names(tp_source), "die")
)
lobstr::tree(
  rapply(valid_tp, function(x) round(x, 2), how = "list")
)
```

Missing destination names within an inner list are auto-added as 0. Omitting an
entire 1st-level element (a whole state) will return an error.


## A note on order of operations

The last step before entering the data into M is to make sure that `valid_tp`,
which is valid in and of itself, corresponds to the specification of `M`.

Well, if we reverse the order of operations, we can make the process more
straightforward by using `valid_tp` to inform some of the inputs `specify_m` needs.
These are:

- `tunnel_lengths`
- `pre_tunnel_states`

Unlike the old `n_cycles` parameter, `tunnel_lengths` cannot be derived
automatically from the transition probability data, because different tunnels
may have different lengths — that is a modelling choice, not something inferable
from TP vector structure. The number of tunnels is not a separate argument; it
is implicit in `length(tunnel_lengths)`.

`pre_tunnel_states` must also be determined by the user, as the structure of the
TP data alone does not distinguish pre-tunnel states from tunnel states. There
are the same number of destination states, and the tp vectors have the same
shapes and names regardless of whether a state is pre-tunnel or not.

It is still most robust to validate the tp data first, and then specify `M`,
because during validation some structural checks — such as requiring at least
one tunnel and one pre-tunnel state — are conducted.

In most oncology settings, only the first state can be a pre-tunnel state
because the transition from e.g., off-treatment in first line to any subsequent
state has an index date of discontinuation event for first line therapy, not
baseline. However, there are feasibly use-cases for multiple pre-tunnel states.
Yet for this example, **we say that only `on_1l` is a pre-tunnel state**.

```{r specify_m_from_valid_tp}
m_spec <- specify_m(
  tunnel_lengths = rep(th_cyc, length(valid_tp) - 1),
  pre_tunnel_states = 1
)
```

## Populating `m1`

The baseline-time transitions will be a list of `r th_cyc` `m1`s in this case, one for
each cycle. Each of these is a matrix with `1` row and `r m_spec$matrix_size`
columns.

We can compile this manually if we want, by doing the following. First, pull out
the probabilities for the health state(s) we need to populate once per cycle:

```{r derive_m1_data}
p_off_1l <- valid_tp$on_1l$off_1l
p_on_2l <- valid_tp$on_1l$on_2l
p_off_2l <- valid_tp$on_1l$off_2l
p_bsc <- valid_tp$on_1l$bsc
p_die <- valid_tp$on_1l$die
m_size <- m_spec$matrix_size
m1_coord <- m_spec$m1_ijx
```

Next, have a look at the structure of `m1_ijx` to see where these need to go:

```{r print_m1_ijx}
print(m_spec$m1_ijx)
```

In this case:

- `1` is staying in `on_1l`
- `2` is discontinuation to `off_1l`
- `52` is starting `on_2l`
- `102` is starting `off_2l`
- `152` is starting `bsc`
- `202` is death

Therefore we can easily make our list like this:

```{r manual_m1_list}
m1_list <- lapply(seq_len(th_cyc), function(model_cycle) {
  m1_coord[, "x"] <- c(
    1 - sum(
      p_off_1l[model_cycle],
      p_on_2l[model_cycle],
      p_off_2l[model_cycle],
      p_bsc[model_cycle],
      p_die[model_cycle]
    ),
    p_off_1l[model_cycle],
    p_on_2l[model_cycle],
    p_off_2l[model_cycle],
    p_bsc[model_cycle],
    p_die[model_cycle]
  )
  m1_coord <- m1_coord[m1_coord[, "x"] > 0, , drop = FALSE]
  Matrix::sparseMatrix(
    i = m1_coord[, "i"],
    j = m1_coord[, "j"],
    x = m1_coord[, "x"],
    dims = c(1, m_size)
  )
})
print(m1_list[[1]])
```

This defines a sparse matrix for each model cycle describing the probabilities
for transition for that model cycle from the pre-tunnel state(s) to all other
states in the model. Note the step to remove 0 probabilities before creating the
sparse matrix. This then removes unnecessary data points and therefore
computation later on. To check all sum to 1 (which they must given the 1 - sum()
in the above!):

```{r assert_m1_sums}
# assert that all m1 rows sum to 1 for all model cycles:
assertthat::assert_that(
  all(unlist(lapply(m1_list, sum)) == 1),
  msg = "m1 rows do not sum to 1"
)
```

In practice, validation steps like these are not necessary because the TPs come
from `valid_tps`, which have already been validated against conditions like
non-negativity and summing to 1. The above is just to prove that it is working
correctly.

To make this a bit less cumbersome, we can use the function `generate_m_list`
to do the above using the validated `tp` list for `m1` and `m2` all in one easy
step. The result will be a list with 2 elements, one for `m1` one for `m2`, with
`m1` being a list of length time horizon, and `m2` being a single sparse matrix.

```{r generate_m_list}
# validate inputs and generate the full matrix m from one function
m <- generate_m_list(
  m_specification = m_spec,
  transition_prob_list = tp_source,
  first_state_name = "on_1l"
)
```

Note that this function automatically filters out probabilities entered as zero.
This then avoids unnecessary computation when running extrapolations, which
become more important in e.g., probabilistic settings. When running a
probabilistic sensitivity analysis, one would need to generate `tp_source` `n`
times, where `n` is the number of iterations, and then generate `m` for each
iteration. Therefore, the efficiency of this step is important to consider.

If you want to check that all rows sum to 1 you can do things like this:
`rowSums(m$m2)`. You *should* see that the first one is zero, and all the others
are exactly 1, like this:

```{r print_rowsums}
rowSums(m$m2)
```


# Extrapolation

From this point it is very simple to generate a Markov trace!

```{r extrapolate_manual}
# output population matrix:
pop <- matrix(0, nrow = m_spec$matrix_size, ncol = th_cyc + 1)
pop[1, 1] <- 1

# get the TPs
m1 <- m$m1
m2 <- m$m2

# extrapolate the model:
for (cyc in seq_len(th_cyc) + 1) {
  pop[, cyc] <-
    as.numeric(pop[seq_len(m_spec$pre_tunnels), cyc - 1] %*% m1[[cyc - 1]]) +
    as.numeric(pop[, cyc - 1] %*% m2)
}
```

Believe it or not, that's the Markov model done, for the most part. The next
step is consolidating information from this matrix `pop`, which has some useful
properties.

## Properties of `pop`

`pop` is an interesting object because you can add up bits of it to get
different bits of information. These are really useful for cost-effectiveness
modelling.

### Overall survival

One minus the last row is overall survival:

```{r plot_os}
plot(1 - pop[nrow(pop), ],
  type = "l", xlab = "Cycle", ylab = "Overall Survival",
  main = "Overall Survival Curve"
)
```

### Tunnel populations in absolute time `t`

Vertical sums of sections of `pop` each of height of time horizon, except for
the pre-tunnel states and death (one row each), are tunnel state populations at
each absolute time point, `t`, so in cycle 2, we have `colSums(pop[2:51, ])[2]`,
which is `r colSums(pop[2:51, ])[2]` i.e., `1` (baseline pop) multiplied by the
probability of discontinuing first-line but not starting second line (`0.1`).  

In English, 0.1 of the cohort is in the first tunnel state in cycle 2.  

Plotting this is easy if you know which rows to add up. In the below we have
entered the row numbers manually to help you see the blocks for yourselves. Note
that the time horizon is 50 cycles in this example:

```{r plot_tunnel_pops}
plot(1 - pop[202, ],
  type = "l",
  col = "black",
  xlab = "Cycle", ylab = "Tunnel 1 Population",
  main = "Tunnel Populations over Time"
)
lines(colSums(pop[2:51, ]), type = "l", col = "grey80")
lines(colSums(pop[52:101, ]), type = "l", col = "red")
lines(colSums(pop[102:151, ]), type = "l", col = "blue")
lines(colSums(pop[152:201, ]), type = "l", col = "green")
```


### Tunnel populations in relative time `d`

This is where `pop` gets more interesting. The sum _across_ rows is the
population from the time point of entering the tunnel, or in other words `d = 0`
for that tunnel. For instance, in the above example, the sum of the 2nd row is
the proportion of the cohort that moved into tunnel 1 at any time in the model.  

If you take the appropriate set of `rowSum` results and divide it by the
first element of the result, this will give you a within-tunnel survival (i.e.,
population starting from 100% remaining inside of the tunnel in question).  

This lets you plot something more familiar, but only if you divide the values by
their first element. Without that, it looks strange:

```{r plot_d_unadjusted}
plot(
  1 - pop[202, ],
  type = "l",
  col = "black",
  xlab = "Cycle",
  ylab = "Within-tunnel survival",
  main = "Tunnel Populations over Time"
)
lines(rowSums(pop[2:51, ]), type = "l", col = "grey80")
lines(rowSums(pop[52:101, ]), type = "l", col = "red")
lines(rowSums(pop[102:151, ]), type = "l", col = "blue")
lines(rowSums(pop[152:201, ]), type = "l", col = "green")
```

But if you adjust it for the proportion of the initial cohort that ever entered
each tunnel, you get something more familiar:

```{r plot_d_adjusted}
plot(
  1 - pop[202, ],
  type = "l",
  col = "black",
  xlab = "Cycle",
  ylab = "Within-tunnel survival",
  main = "Tunnel Populations over Time"
)
lines(rowSums(pop[2:51, ]) / sum(pop[2, ]), type = "l", col = "grey80")
lines(rowSums(pop[52:101, ]) / sum(pop[52, ]), type = "l", col = "red")
lines(rowSums(pop[102:151, ]) / sum(pop[102, ]), type = "l", col = "blue")
lines(rowSums(pop[152:201, ]) / sum(pop[152, ]), type = "l", col = "green")
```

This is the **within-tunnel survival curve** for each tunnel, which is much more
intuitive, and is more likely to be comparable to some external data for
external validation purposes. For instance, if you have progression-free
survival for patients entering 3rd line therapy, you need the line to start from
100%, not the proportion of patients in your model that entered that state.

However, **one critical warning** here is that the source Kaplan-Meier used to
inform the transition probabilities for these tunnel states **will not align**
with these curves. This is because the appropriate endpoint for informing the
transition probabilities needs to have carefully designed censoring rules to
isolate the hazard of a specific transition. Therefore, these curves, which also
include exiting the state for _any_ reason (not just one specific event) should
be compared to Kaplan-Meier which **does not e.g., censor for death**. In other
words, these extrapolations should be compared to within-state survival, not
e.g., time to progression. 

An example of appropriate Kaplan-Meier setup for e.g., a 3rd line therapy would
be (all from index date of initiating 3rd line):

- Kaplan-Meier **informing transition probabilities**
  - Time to initiation of 4th line therapy censoring for death and patient no treatment decision
  - time to patient no treatment decision censoring for death and 4th line initiation
  - Time to death censoring for 4th line initiation and patient no treatment decision
- Kaplan-Meier **for validation**
  - 4th line free survival
  - Patient no treatment decision free survival
  - Overall survival

As you can see this can be quite the task for a statistics department, as they
have to carefully handle their dataset to correctly derive all the event and
censor definitions and implement them correctly. However, when done correctly,
the final traces (starting from 100% survival) should match up with the
validation Kaplan-Meier curves closely. It is possible that some straying will
be visible, especially towards the tail of the available follow up. This is
often because the data is being spread quite thin due to all the censoring on
the individual endpoints. However, if the trial is quite large, this will be
mostly mitigated.

`treatseqr` is about using these probabilities and not about deriving them, so
we will limit our explanation here and assume that users know what they are
doing with trial statistics.

As a last word on the matter: as long careful consideration is made around the
data these lines should be compared to first, these plots can be valuable
sources of both internal and external validation.

## Functions to generate outputs

The full matrix is the most useful object to return for `pop`. However,
interpreting it then requires manual steps to add things up in ways that
typically will always be needed for cost-effectiveness modelling in health
economics. Yet, individual preferences will always be a factor, so we have
broken down the steps into separate simple functions to do this for you.

### Extrapolation

The extrapolation function is trivial, but still useful to have. This function
extrapolates the model, returning the full trace for you to do with as you
please:

```{r extrapolate}
trace <- extrapolate_treatseqr(m = m, spec = m_spec)
colSums(trace)
```


### Consolidation

After the trace runs, we need to consolidate the results into useful outputs.
This usually includes both the vertical and horizontal sums (see above). The
function `consolidate_treatseqr_trace` does this for you, returning a list
containing both.

Interestingly, the bounds of the states within `trace` always correspond to the
numbers in `m1_ijx` in the `j` column, which simplifies the process of adding
things up nicely. We just use a simple algorithm leveraging information stored
in `m_spec` to figure out which pairings of values in `m1_ijx[, "j"]` correspond
to which state:

- `pre_tunnels`: The first `pre_tunnels` rows in `trace` are 1 row/column each
and these are states which have a basis in time of `t` not `d`, so are simply
the appropriate row (e.g., `trace[1, ]` is the first health state over time).
- Tunnels: each tunnel has rows/columns corresponding to specific sizing, and
these can be summed to produce what we need.
- Death: The last row/column is death, which is absorbing.


```{r consolidate}
trace <- consolidate_treatseqr_trace(
  full_trace = trace,
  spec = m_spec,
  state_names = m$state_names
)

# expected time in state in model cycles:
colSums(trace$t)
```

These can be used for easy plotting, using `matplot` from `base`.

```{r matplot_t}
graphics::matplot(trace$t, type = "l")
graphics::legend(
  "topright",
  legend = colnames(trace$t),
  col = seq_len(ncol(trace$t)),
  lty = 1
)
```



Similarly, we can look at the proportional survival in time in state `d`.

```{r matplot_d}
graphics::matplot(trace$d, type = "l")
graphics::legend(
  "topright",
  legend = colnames(trace$d),
  col = seq_len(ncol(trace$d)),
  lty = 1
)
```
