---
title: "What if it's all tunnels?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{What if it's all tunnels?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(Matrix)
library(heseqmat)
library(microbenchmark)
```

We wonder whether it might be more efficient to use tunnel states for *all*
time-dependency in the model. This is the case if it is impossible to return to
the initial state and there is no other need to depend on wall time. There will
likely be examples where this assumption is unreasonable, e.g., because wall
time is a measure of *age*, which is frequently a predictor of mortality.

We will use a familiar example:

```{r define_example}
th_cyc <- 50

# some simple time varying transition probabilities. in reality these will come
# from hazard extrapolations e.g. from `flexsurv`
valid_tp <- list(
  on_1l = list(
    off_1l = seq(0.1, 0.05, length.out = th_cyc), # Example: linearly decreasing
    on_2l = seq(0.05, 0.1, length.out = th_cyc), # Example: linearly increasing
    off_2l = rep(0, th_cyc),
    bsc = rep(0, th_cyc),
    die = rep(0.01, th_cyc)
  ),
  off_1l = list(
    on_2l = seq(0.08, 0.02, length.out = th_cyc),
    off_2l = seq(0.01, 0.03, length.out = th_cyc),
    bsc = rep(0.02, th_cyc),
    die = seq(0.05, 0.1, length.out = th_cyc)
  ),
  on_2l = list(
    off_2l = seq(0.05, 0.02, length.out = th_cyc),
    bsc = rep(0.01, th_cyc),
    die = seq(0.1, 0.15, length.out = th_cyc)
  ),
  off_2l = list(
    bsc = seq(0.075, 0.05, length.out = th_cyc),
    die = seq(0.15, 0.2, length.out = th_cyc)
  ),
  bsc = list(
    die = seq(0.2, 0.3, length.out = th_cyc)
  )
) |>
  validate_tp_source()

m_spec <- specify_m(
  n_cycles = th_cyc,
  n_tunnels = length(valid_tp) - 1,
  pre_tunnel_states = 1
)

# validate inputs and generate the full matrix m from one function
m <- generate_m_list(
  m_specification = m_spec,
  transition_prob_list = valid_tp
)
```

Benchmark the existing approach:

```{r benchmark-existing}
microbenchmark(extrapolate_heseqmat(m = m, spec = m_spec), times = 1000)
```

Now let's try only using tunnel states

```{r setup-only-tunnel-states}
tunnel_starts <- 1L + th_cyc * 0:(length(valid_tp) - 1)
death_state <- th_cyc * length(valid_tp) + 1
ii <- c(
  (tunnel_starts[1]):(tunnel_starts[2]-1L), # on_1L -> on_1L
  (tunnel_starts[1]):(tunnel_starts[2]-1L), # on_1L -> off_1L
  (tunnel_starts[1]):(tunnel_starts[2]-1L), # on_1L -> on_2L
  (tunnel_starts[1]):(tunnel_starts[2]-1L), # on_1L -> off_2L
  (tunnel_starts[1]):(tunnel_starts[2]-1L), # on_1L -> bsc
  (tunnel_starts[1]):(tunnel_starts[2]-1L), # on_1L -> die
  (tunnel_starts[2]):(tunnel_starts[3]-1L), # off_1L -> off_1L
  (tunnel_starts[2]):(tunnel_starts[3]-1L), # off_1L -> on_2L
  (tunnel_starts[2]):(tunnel_starts[3]-1L), # off_1L -> off_2L
  (tunnel_starts[2]):(tunnel_starts[3]-1L), # off_1L -> bsc
  (tunnel_starts[2]):(tunnel_starts[3]-1L), # off_1L -> die
  (tunnel_starts[3]):(tunnel_starts[4]-1L), # on_2L -> on_2L
  (tunnel_starts[3]):(tunnel_starts[4]-1L), # on_2L -> off_2L
  (tunnel_starts[3]):(tunnel_starts[4]-1L), # on_2L -> bsc
  (tunnel_starts[3]):(tunnel_starts[4]-1L), # on_2L -> die
  (tunnel_starts[4]):(tunnel_starts[5]-1L), # off_2L -> off_2L
  (tunnel_starts[4]):(tunnel_starts[5]-1L), # off_2L -> bsc
  (tunnel_starts[4]):(tunnel_starts[5]-1L), # off_2L -> die
  (tunnel_starts[5]):(death_state-1L),      # bsc -> bsc
  (tunnel_starts[5]):(death_state-1L),      # bsc -> die
  death_state                               # die -> die
)
jj <- c(
  (tunnel_starts[1]+1L):(tunnel_starts[2]-1L), tunnel_starts[2]-1L, # on_1L -> on_1L
  rep(tunnel_starts[2], th_cyc),                                    # on_1L -> off_1L
  rep(tunnel_starts[3], th_cyc),                                    # on_1L -> on_2L
  rep(tunnel_starts[4], th_cyc),                                    # on_1L -> off_2L
  rep(tunnel_starts[5], th_cyc),                                    # on_1L -> bsc
  rep(death_state, th_cyc),                                         # on_1L -> die
  (tunnel_starts[2]+1L):(tunnel_starts[3]-1L), tunnel_starts[3]-1L, # off_1L -> off_1L
  rep(tunnel_starts[3], th_cyc),                                    # off_1L -> on_2L
  rep(tunnel_starts[4], th_cyc),                                    # off_1L -> off_2L
  rep(tunnel_starts[5], th_cyc),                                    # off_1L -> bsc
  rep(death_state, th_cyc),                                         # off_1L -> die
  (tunnel_starts[3]+1L):(tunnel_starts[4]-1L), tunnel_starts[4]-1L, # on_2L -> on_2L
  rep(tunnel_starts[4], th_cyc),                                    # on_2L -> off_2L
  rep(tunnel_starts[5], th_cyc),                                    # on_2L -> bsc
  rep(death_state, th_cyc),                                         # on_2L -> die
  (tunnel_starts[4]+1L):(tunnel_starts[5]-1L), tunnel_starts[5]-1L, # off_2L -> off_2L
  rep(tunnel_starts[5], th_cyc),                                    # off_2L -> bsc
  rep(death_state, th_cyc),                                         # off_2L -> die
  (tunnel_starts[5]+1L):(death_state-1L), death_state-1L,           # bsc -> bsc
  rep(death_state, th_cyc),                                         # bsc -> die
  death_state                                                       # die -> die
)
xx <- c(
  1 - Reduce(`+`, valid_tp$on_1l),  # on_1L -> on_1L
  valid_tp$on_1l$off_1l,            # on_1L -> off_1L
  valid_tp$on_1l$on_2l,             # on_1L -> on_2L
  valid_tp$on_1l$off_2l,            # on_1L -> off_2L
  valid_tp$on_1l$bsc,               # on_1L -> bsc
  valid_tp$on_1l$die,               # on_1L -> die
  1 - Reduce(`+`, valid_tp$off_1l), # off_1L -> off_1L
  valid_tp$off_1l$on_2l,            # off_1L -> on_2L
  valid_tp$off_1l$off_2l,           # off_1L -> off_2L
  valid_tp$off_1l$bsc,              # off_1L -> bsc
  valid_tp$off_1l$die,              # off_1L -> die
  1 - Reduce(`+`, valid_tp$on_2l),  # on_2L -> on_2L
  valid_tp$on_2l$off_2l,            # on_2L -> off_2L
  valid_tp$on_2l$bsc,               # on_2L -> bsc
  valid_tp$on_2l$die,               # on_2L -> die
  1 - Reduce(`+`, valid_tp$off_2l), # off_2L -> on_2L
  valid_tp$off_2l$bsc,              # off_2L -> bsc
  valid_tp$off_2l$die,              # off_2L -> die
  1 - valid_tp$bsc$die,             # bsc -> bsc
  valid_tp$bsc$die,                 # bsc -> die
  1                                 # die -> die
)
m_all_tunnels <- Matrix::sparseMatrix(
  i = ii, j = jj, x = xx, dims = rep(death_state, 2)
)
```

```{r benchmark-only-tunnel-states}
extrapolate_all_tunnels <- function(m_all_tunnels) {
  pop <- matrix(0, nrow = death_state, ncol = th_cyc + 1)
  pop[1,1] <- 1
  for (cyc in seq_len(th_cyc) + 1)
    pop[,cyc] <- as.numeric(pop[,cyc-1] %*% m_all_tunnels)
  pop
}
microbenchmark(extrapolate_all_tunnels(m_all_tunnels), times = 1000)
```

And let's check they actually give the same results...

```{r check-equal}

trace_existing <- extrapolate_heseqmat(m, m_spec)
trace_all_tunnels <- extrapolate_all_tunnels(m_all_tunnels)

sumtrace_existing <- rbind(
  trace_existing[1,],
  colSums(trace_existing[2:51,]),
  colSums(trace_existing[52:101,]),
  colSums(trace_existing[102:151,]),
  colSums(trace_existing[152:202,]),
  trace_existing[203,]
)

sumtrace_all_tunnels <- rbind(
  colSums(trace_all_tunnels[1:50,]),
  colSums(trace_all_tunnels[51:100,]),
  colSums(trace_all_tunnels[101:150,]),
  colSums(trace_all_tunnels[151:200,]),
  colSums(trace_all_tunnels[201:250,]),
  trace_all_tunnels[251,]
)

max(abs(sumtrace_existing - sumtrace_all_tunnels))

```

They are giving the same result, except for numerical precision.

So we should consider relaxing the condition in `specify_m()` that
`pre_tunnel_states > 0`, and allow it to produce an empty `m1`, and optimise
accordingly when this is found.
